---
title: "Getting started with artoo"
vignette: >
  %\VignetteIndexEntry{Getting started with artoo}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| include: false
library(artoo)
```

`artoo` carries clinical-trial datasets losslessly across SAS XPORT,
CDISC Dataset-JSON, NDJSON, Apache Parquet, and RDS through one canonical,
CDISC-shaped metadata model. This vignette walks the whole workflow once,
start to finish: **spec → apply → inspect → write → read back**.

## 1. Get a spec

A `artoo_spec` is the canonical description of your datasets: variables,
CDISC data types, lengths, labels, controlled-terminology codelists, and
sort keys — always for exactly **one** CDISC standard. `read_spec()` reads
one from Define-XML, a Pinnacle 21 workbook, or artoo's native JSON;
`artoo_spec()` assembles one from metadata frames.

The package bundles ready-made specs built from the official CDISC
Define-XML 2.1 release examples — `adam_spec` for ADaM (ADSL, ADAE) and
`sdtm_spec` for SDTM (DM, VS, TS, SUPPDM). Each also ships as a P21
workbook you can open in Excel:

```{r}
adam_spec

p21 <- system.file("extdata", "adam-spec.xlsx", package = "artoo")
identical(spec_standard(read_spec(p21)), spec_standard(adam_spec))
```

Because `read_spec()` and `write_spec()` are inverses on each format,
format conversion is one composition — Define-XML in, P21 workbook out:

```r
read_spec("define.xml") |> write_spec("spec.xlsx")
```

## 2. Apply the spec

`apply_spec()` is the conform pipeline: it scaffolds the variables the spec
declares (typed `NA`), coerces each column to its CDISC data type, orders
the columns, sorts by the dataset keys, and stamps the result with its
metadata. The input is never mutated, no column is ever dropped, and a
coercion that would damage values aborts before it runs.

```{r}
adsl <- apply_spec(cdisc_adsl, adam_spec, "ADSL")
```

The conformance findings ride along on the result — read them back as a
frame with `conformance()`:

```{r}
nrow(conformance(adsl))
```

The pipeline is standard-neutral: an SDTM domain conforms identically —
only the spec and the dataset change. The full SDTM walkthrough is its own
[task guide](articles/sdtm-build.html).

```{r}
dm <- apply_spec(cdisc_dm, sdtm_spec, "DM")
nrow(conformance(dm))
```

## 3. Inspect the columns

`columns()` is the quick look a SAS programmer expects from
`PROC CONTENTS`: one row per variable with position, type, length, format,
label, and the CDISC key sequence. It works on a conformed frame, any
plain data frame, or a file path:

```{r}
columns(adsl)
```

## 4. Write to any format — losslessly

Every writer carries the full metadata model, so the write is lossless by
construction. The writers return their input invisibly, so one conformed
frame fans out to every deliverable:

```{r}
xpt <- tempfile(fileext = ".xpt")
json <- tempfile(fileext = ".json")

adsl |>
  write_xpt(xpt) |>
  write_json(json)
```

Any file converts to any other without re-applying the spec — the
metadata travels inside (or beside) the container:

```{r}
parquet <- tempfile(fileext = ".parquet")
write_parquet(read_json(json), parquet)
```

## 5. Read back, intact

Reading restores the values, the R classes (dates as `Date`, times as
`hms`), the labels, and the metadata — identically from every format:

```{r}
back <- read_json(json)
get_meta(back)@dataset$records
columns(back)
```

One honest caveat: the XPORT byte layout stores only name, label, length,
and formats, so `columns()` on an `.xpt` path shows a blank `Key` — the
key sequence (like codelist references) rides the metadata-carrying
formats and the in-session frame, never the 1980s transport bytes.

That round-trip identity is the whole point: what you submit is what you
archived is what you analysed.

## Where to next

- The task guides: [an end-to-end ADaM build](articles/adam-build.html),
  [an end-to-end SDTM build](articles/sdtm-build.html),
  [any-to-any conversion](articles/convert.html),
  [dates, times, and `--DTC`](articles/dates.html),
  [validation & qualification](articles/validation.html), and
  [common errors](articles/errors.html).
- `?apply_spec`, `?check_spec`, `?decode_column` — the conform surface.
- `?read_spec`, `?write_spec`, `?spec_variables` — the spec surface.
- `?read_dataset` — generic I/O and the per-format wrappers.
- `?adam_spec` — the bundled demo specs and datasets.
