---
title: "Get started with artoo"
vignette: >
  %\VignetteIndexEntry{Get started with artoo}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| include: false
library(artoo)
```

`artoo` carries clinical-trial datasets losslessly across SAS XPORT,
CDISC Dataset-JSON, NDJSON, Apache Parquet, and RDS through one canonical,
CDISC-shaped metadata model. The whole package is a single loop: a **spec**
plus **data** go through `apply_spec()`, write to a **file**, and read back
**identical**. That loop is the lossless guarantee.

::: {.rt-figure}
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 1040 250" width="100%" role="img" aria-label="The artoo lossless round-trip: spec plus data go through apply_spec (scaffold, coerce, order, sort, stamp), then write to a file, then read back to identical data; set_type and check_spec fix and inspect the spec, and the whole loop is lossless, so what you read back equals what you wrote.">
  <defs>
    <marker id="rt-arrow" markerWidth="10" markerHeight="10" refX="7" refY="3" orient="auto" markerUnits="userSpaceOnUse">
      <path d="M0,0 L7,3 L0,6 Z" fill="#94a3b8"/>
    </marker>
    <marker id="rt-arrow-blue" markerWidth="11" markerHeight="11" refX="7.5" refY="3.2" orient="auto" markerUnits="userSpaceOnUse">
      <path d="M0,0 L7.5,3.2 L0,6.4 Z" fill="#3b82f6"/>
    </marker>
  </defs>
  <rect x="1" y="1" width="1038" height="248" fill="#ffffff" stroke="#efefef" stroke-width="2"/>
  <text x="99" y="32" text-anchor="middle" font-family="system-ui, -apple-system, 'Segoe UI', sans-serif" font-size="10.5" fill="#64748b">fix &#183; inspect the spec</text>
  <text x="99" y="51" text-anchor="middle" font-family="ui-monospace, SFMono-Regular, Menlo, monospace" font-size="12" fill="#1e293b">set_type() &#183; check_spec()</text>
  <line x1="99" y1="60" x2="99" y2="83" stroke="#94a3b8" stroke-width="1.6" marker-end="url(#rt-arrow)"/>
  <g stroke="#94a3b8" stroke-width="1.8">
    <line x1="176" y1="112" x2="207" y2="112" marker-end="url(#rt-arrow)"/>
    <line x1="382" y1="112" x2="413" y2="112" marker-end="url(#rt-arrow)"/>
    <line x1="538" y1="112" x2="569" y2="112" marker-end="url(#rt-arrow)"/>
    <line x1="666" y1="112" x2="697" y2="112" marker-end="url(#rt-arrow)"/>
    <line x1="822" y1="112" x2="853" y2="112" marker-end="url(#rt-arrow)"/>
  </g>
  <rect x="24" y="86" width="150" height="52" fill="#f9f9f9" stroke="#e5e7eb" stroke-width="1.5"/>
  <text x="99" y="112" text-anchor="middle" dominant-baseline="central" font-family="ui-monospace, SFMono-Regular, Menlo, monospace" font-size="14" fill="#1e293b">spec + data</text>
  <rect x="210" y="86" width="170" height="52" fill="#3b82f6" stroke="#2563eb" stroke-width="1.5"/>
  <text x="295" y="112" text-anchor="middle" dominant-baseline="central" font-family="ui-monospace, SFMono-Regular, Menlo, monospace" font-size="14.5" font-weight="600" fill="#ffffff">apply_spec()</text>
  <text x="295" y="160" text-anchor="middle" font-family="system-ui, -apple-system, 'Segoe UI', sans-serif" font-size="11" fill="#64748b">scaffold &#183; coerce &#183; order &#183; sort &#183; stamp</text>
  <rect x="416" y="86" width="120" height="52" fill="#f9f9f9" stroke="#e5e7eb" stroke-width="1.5"/>
  <text x="476" y="112" text-anchor="middle" dominant-baseline="central" font-family="ui-monospace, SFMono-Regular, Menlo, monospace" font-size="14" fill="#1e293b">write_*()</text>
  <rect x="572" y="86" width="92" height="52" fill="#eef3fb" stroke="#c7dbff" stroke-width="1.5"/>
  <text x="618" y="112" text-anchor="middle" dominant-baseline="central" font-family="ui-monospace, SFMono-Regular, Menlo, monospace" font-size="14" fill="#1e3a5f">file</text>
  <rect x="700" y="86" width="120" height="52" fill="#f9f9f9" stroke="#e5e7eb" stroke-width="1.5"/>
  <text x="760" y="112" text-anchor="middle" dominant-baseline="central" font-family="ui-monospace, SFMono-Regular, Menlo, monospace" font-size="14" fill="#1e293b">read_*()</text>
  <rect x="856" y="86" width="160" height="52" fill="#f9f9f9" stroke="#e5e7eb" stroke-width="1.5"/>
  <text x="936" y="112" text-anchor="middle" dominant-baseline="central" font-family="ui-monospace, SFMono-Regular, Menlo, monospace" font-size="14" fill="#1e293b">identical data</text>
  <path d="M936,140 V202 H99 V143" fill="none" stroke="#3b82f6" stroke-width="1.8" stroke-dasharray="6 5" marker-end="url(#rt-arrow-blue)"/>
  <text x="517" y="228" text-anchor="middle" font-family="system-ui, -apple-system, 'Segoe UI', sans-serif" font-size="13" font-weight="600" fill="#3b82f6">lossless round-trip &#8212; what you read back equals what you wrote</text>
</svg>
:::

This vignette walks that loop once, start to finish: **spec &rarr; apply
&rarr; inspect &rarr; write &rarr; read back**, all on the bundled demo data.

## 1. Get a spec

A `artoo_spec` is the canonical description of your datasets: variables,
CDISC data types, lengths, labels, controlled-terminology codelists, and
sort keys &mdash; always for exactly **one** CDISC standard. `read_spec()` reads
one from Define-XML, a Pinnacle 21 workbook, or artoo's native JSON;
`artoo_spec()` assembles one from metadata frames.

The package bundles ready-made specs built from the official CDISC
Define-XML 2.1 release examples &mdash; `adam_spec` for ADaM (ADSL, ADAE) and
`sdtm_spec` for SDTM (DM, VS, TS, SUPPDM). Each also ships as a P21
workbook you can open in Excel:

```{r}
adam_spec

p21 <- system.file("extdata", "adam-spec.xlsx", package = "artoo")
identical(spec_standard(read_spec(p21)), spec_standard(adam_spec))
```

Because `read_spec()` and `write_spec()` are inverses on each format,
format conversion is one composition &mdash; Define-XML in, P21 workbook out:

```r
read_spec("define.xml") |> write_spec("spec.xlsx")
```

## 2. Apply the spec

`apply_spec()` is the conform pipeline: it coerces each column to its CDISC
data type, orders the columns, sorts by the dataset keys, and stamps the
result with its metadata. A variable the spec declares but the data lacks is
reported, never fabricated as an empty column. The input is never mutated, no
column is ever dropped, and a coercion that would damage values aborts before
it runs &mdash; with two honest one-line exits: keep the wider source type with
`apply_spec(..., on_coercion_loss = "keep")`, or retype the spec with
`set_type()`.

```{r}
adsl <- apply_spec(cdisc_adsl, adam_spec, "ADSL")
```

The conformance findings ride along on the result &mdash; read them back as a
frame with `conformance()`:

```{r}
nrow(conformance(adsl))
```

The pipeline is standard-neutral: an SDTM domain conforms identically &mdash;
only the spec and the dataset change.

```{r}
dm <- apply_spec(cdisc_dm, sdtm_spec, "DM")
nrow(conformance(dm))
```

## 3. Inspect the columns

`columns()` is the quick look a SAS programmer expects from
`PROC CONTENTS`: one row per variable with position, type, length, format,
label, and the CDISC key sequence. It works on a conformed frame, any
plain data frame, or a file path:

```{r}
columns(adsl)
```

## 4. Write to any format — losslessly

Every writer carries the full metadata model, so the write is lossless by
construction. The writers return their input invisibly, so one conformed
frame fans out to every deliverable:

```{r}
xpt <- tempfile(fileext = ".xpt")
json <- tempfile(fileext = ".json")

adsl |>
  write_xpt(xpt) |>
  write_json(json)
```

Any file converts to any other without re-applying the spec &mdash; the
metadata travels inside (or beside) the container:

```{r}
parquet <- tempfile(fileext = ".parquet")
write_parquet(read_json(json), parquet)
```

## 5. Read back, intact

Reading restores the values, the R classes (dates as `Date`, times as
`hms`), the labels, and the metadata &mdash; identically from every format:

```{r}
back <- read_json(json)
get_meta(back)@dataset$records
columns(back)
```

One honest caveat: the XPORT byte layout stores only name, label, length,
and formats, so `columns()` on an `.xpt` path shows a blank `Key` &mdash; the
key sequence (like codelist references) rides the metadata-carrying
formats and the in-session frame, never the 1980s transport bytes.

That round-trip identity is the whole point: what you submit is what you
archived is what you analysed.

## Where to next

- The task guides: [any-to-any conversion](https://vthanik.github.io/artoo/articles/convert.html)
  and [validation & qualification](https://vthanik.github.io/artoo/articles/validation.html).
- `?apply_spec`, `?check_spec`, `?decode_column` — the conform surface.
- `?read_spec`, `?write_spec`, `?spec_variables` — the spec surface.
- `?read_dataset` — generic I/O and the per-format wrappers.
- `?adam_spec` — the bundled demo specs and datasets.
```
