---
title: "From spec to submission dataset"
vignette: >
  %\VignetteIndexEntry{From spec to submission dataset}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| include: false
library(artoo)
```

`artoo` turns a raw analysis frame and a dataset specification into a
submission-ready dataset, then writes it to any clinical exchange format
without losing a single piece of metadata. This vignette walks the core
workflow on the bundled CDISC pilot data: **spec → apply → check → write**.

## 1. Build a spec

A `artoo_spec` is the canonical, CDISC-shaped description of your datasets:
the variables, their data types and lengths, controlled-terminology
codelists, sort keys, and labels. You can read one from a Define-XML or
Pinnacle 21 workbook with `read_spec()`, or assemble it from metadata
frames with `artoo_spec()`.

The package ships the pilot metadata as `cdisc_datasets`, `cdisc_variables`,
and `cdisc_codelists`:

```{r}
spec <- artoo_spec(
  cdisc_datasets,
  cdisc_variables,
  codelists = cdisc_codelists
)
spec
```

## 2. Apply the spec

`apply_spec()` is the conformance pipeline. It scaffolds the columns the
spec declares, coerces each to its CDISC data type, orders the columns,
sorts by the key sequence, and stamps the result with its metadata. The
raw `cdisc_adsl` frame goes in; a conformed ADSL comes out.

```{r}
adsl <- apply_spec(cdisc_adsl, spec, "ADSL")
dim(adsl)
```

The conformed frame carries its `artoo_meta`, which `get_meta()` reads back.
This is the single source of truth that every codec preserves:

```{r}
meta <- get_meta(adsl)
meta@dataset$records
```

## 3. Check conformance

`check_spec()` returns a tidy findings frame: every place the data departs
from the spec (unknown columns, out-of-codelist values, length overflows,
type mismatches). A clean dataset returns zero rows.

```{r}
findings <- check_spec(adsl, spec, "ADSL")
nrow(findings)
```

`validate_spec()` is the complementary check on the *spec itself* —
intrinsic integrity, independent of any data:

```{r}
validate_spec(spec)
```

## 4. Write a submission dataset

With a conformed, checked frame in hand, write it to SAS XPORT v5 — the
format the FDA expects:

```{r}
path <- tempfile(fileext = ".xpt")
write_xpt(adsl, path)
```

Read it straight back and the metadata is intact — labels and SAS formats
ride along on the columns, the records count matches, the types are
preserved:

```{r}
back <- read_xpt(path)
get_meta(back)@dataset$records
```

That round-trip losslessness is the whole point: see
`vignette("one-dataset-every-format")` for converting one conformed dataset
across XPT, Dataset-JSON, Parquet, and RDS interchangeably.
