---
title: "One dataset, every format"
vignette: >
  %\VignetteIndexEntry{One dataset, every format}
  %\VignetteEngine{quarto::format}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| include: false
library(artoo)
```

A single conformed dataset can be written to any clinical exchange format
and read back without losing metadata. Because every codec carries the same
canonical `artoo_meta`, conversion between any two formats is lossless *by
construction* — not by best effort. This vignette demonstrates that with one
ADSL dataset and all four supported formats.

## The formats

`artoo_formats()` lists what `artoo` reads and writes:

```{r}
artoo_formats()
```

SAS XPORT (XPT) for FDA submission, CDISC Dataset-JSON v1.1 for modern
interchange, Apache Parquet for analytics, and RDS for fast R-native
storage.

## Start from a conformed dataset

We conform ADSL exactly as in `vignette("from-spec-to-submission")`:

```{r}
spec <- artoo_spec(cdisc_datasets, cdisc_variables, codelists = cdisc_codelists)
adsl <- apply_spec(cdisc_adsl, spec, "ADSL")
dim(adsl)
```

## Write to every format

The per-format writers all take `(x, path)`; the generic `write_dataset()`
dispatches on the file extension. Here we fan the same frame out to four
files:

```{r}
paths <- list(
  xpt     = tempfile(fileext = ".xpt"),
  json    = tempfile(fileext = ".json"),
  parquet = tempfile(fileext = ".parquet"),
  rds     = tempfile(fileext = ".rds")
)

write_xpt(adsl, paths$xpt)
write_json(adsl, paths$json)
write_parquet(adsl, paths$parquet)
write_rds(adsl, paths$rds)
```

## Read each back and compare

Reading any of the four returns the same rows and the same metadata. The
record count is identical across all formats:

```{r}
back <- list(
  xpt     = read_xpt(paths$xpt),
  json    = read_json(paths$json),
  parquet = read_parquet(paths$parquet),
  rds     = read_rds(paths$rds)
)

vapply(back, nrow, integer(1))
```

The metadata survives every hop. The dataset label, declared on the spec,
comes back identically whether it travelled through XPT's 40-byte header
field or Dataset-JSON's `label` key:

```{r}
vapply(back, function(d) get_meta(d)@dataset$label, character(1))
```

## Any-to-any conversion

Because the metadata is intact on read, a file in one format becomes a file
in another with no spec re-application — read here, write there:

```{r}
# Dataset-JSON on disk -> a submission XPT, metadata carried through
from_json <- read_json(paths$json)
out_xpt <- tempfile(fileext = ".xpt")
write_xpt(from_json, out_xpt)

identical(
  get_meta(read_xpt(out_xpt))@dataset$records,
  get_meta(adsl)@dataset$records
)
```

One canonical metadata model in the middle is what makes every pair of
formats a lossless conversion.
