---
title: "An end-to-end ADaM build"
format: html
---

```{r}
#| include: false
library(artoo)
```

This article walks the everyday ADaM programming loop with artoo doing the
carriage: load the spec, derive, translate through codelists, conform, and
fan the result out to every deliverable format. Everything runs on the
bundled demo data (`cdisc_adsl`, `adam_spec` — built reproducibly from the
official CDISC Define-XML 2.1 example and the PHUSE Test Data Factory).

## 1. The spec is the contract

`adam_spec` describes ADSL and ADAE for ADaMIG 1.1 — variables, CDISC data
types, lengths, labels, codelists, and keys:

```{r}
adam_spec
```

The accessors return plain data frames, so the spec slots into ordinary
base R work:

```{r}
vars <- spec_variables(adam_spec, "ADSL")
head(vars[, c("variable", "label", "data_type", "codelist_id")])
```

## 2. Derive, with temporaries

A real derivation program accumulates working columns the spec never
declares. That is fine — they are "extras", and you decide their fate at
conform time:

```{r}
adsl <- cdisc_adsl
adsl$AGEGR1_TMP <- cut(
  adsl$AGE,
  breaks = c(-Inf, 64, 74, Inf),
  labels = c("<65", "65-74", ">=75")
)
adsl$AGEGR1 <- as.character(adsl$AGEGR1_TMP)
```

## 3. Codelists: check membership, translate values

The codelist surface is reference-keyed: each variable's row in the spec
names its `codelist_id`, and `spec_codelists()` returns the terms:

```{r}
head(spec_codelists(adam_spec))
```

`decode_column()` is the single-variable translator (code to decode or
decode to code — the `RACEN`-from-`RACE` shape); `apply_spec()` itself
never translates values, so submission values stay untouched unless you
ask.

## 4. Conform: scaffold, coerce, order, sort, stamp

`apply_spec()` runs the fixed pipeline and stamps the CDISC metadata.
`extra = "drop"` trims the derivation temporaries to exactly the spec's
columns — announced, and recorded by the `extra_variable` finding, so the
drop is never silent:

```{r}
conformed <- apply_spec(adsl, adam_spec, "ADSL", extra = "drop")
"AGEGR1_TMP" %in% names(conformed)
```

Two boundaries worth knowing, because they are deliberate:

- **`conformance =` disposes of findings only.** Pipeline errors — above
  all a lossy coercion, where the spec says `integer` but the data carries
  fractions — abort under *every* setting (`artoo_error_type`). The fix is
  retyping the spec variable to `"float"`/`"decimal"`, never accepting
  silent truncation. The condition carries the offenders as data in
  `cnd$variables`.
- **Sorting matches SAS.** The default `na_position = "first"` orders
  missing key values before present ones, exactly as SAS `PROC SORT` does
  (and as FDA submission datasets are sorted). Set `"last"` only when your
  comparison target is R's `order()` / pandas / Polars.

## 5. Fan out to every deliverable

The writers return their input invisibly, so one conformed frame flows to
every format in a single pipe — losslessly, because each codec carries the
same metadata:

```{r}
xpt <- tempfile(fileext = ".xpt")
json <- tempfile(fileext = ".json")
conformed |>
  write_xpt(xpt) |>
  write_json(json)
columns(json)
```

## Where to next

- [Any-to-any conversion](convert.html) — moving between the formats.
- [Dates, times, and `--DTC`](dates.html) — how temporal values travel.
- [Common errors](errors.html) — every condition, its trigger, its fix.
