---
title: "Common errors and how to fix them"
format: html
---

```{r}
#| include: false
library(vport)
spec <- vport_spec(cdisc_datasets, cdisc_variables, codelists = cdisc_codelists)
```

Every vport error carries a class (`vport_error_<kind>`) you can catch
programmatically, and a message that tries to name the fix. This page
expands the ones users meet most, with the situation that triggers each
and the way out.

## "Data does not conform to the spec" (`vport_error_conformance`)

`apply_spec(..., conformance = "abort")` found error-severity findings.
The findings are data, not prose — inspect them:

```{r}
dm <- suppressWarnings(apply_spec(cdisc_dm, spec, "DM"))
head(as.data.frame(conformance(dm)), 3)
```

Fix the data or the spec per finding; silence a dimension deliberately
with `vport_checks()`; or run `conformance = "off"` while iterating.

## "Coercion to the spec dataTypes would lose data" (`vport_error_type`)

The spec says `integer` but the data holds fractions (or values beyond
R's 32-bit range). vport aborts *before* truncating — a height of 162.6
silently becoming 162 is a data-integrity event:

```{r}
#| error: true
frac_spec <- vport_spec(
  data.frame(dataset = "ADVS"),
  data.frame(
    dataset = "ADVS", variable = c("USUBJID", "HEIGHTBL"),
    data_type = c("string", "integer")
  )
)
advs <- data.frame(USUBJID = "01-001", HEIGHTBL = 162.6)
apply_spec(advs, frac_spec, "ADVS", conformance = "off")
```

Almost always the spec is wrong: `HEIGHTBL` should be `float` (or
`decimal`). If the truncation is genuinely intended, opt in with
`on_lossy = "warn"`. `check_spec()` flags the same condition before any
coercion as the `integer_fraction` dimension.

## "Values in `VAR` are not in codelist" (`vport_error_codelist`)

A decode (or `decode_column()`) hit a value the codelist does not
define:

```{r}
#| error: true
bad <- cdisc_dm
bad$SEX[1] <- "X"
apply_spec(bad, spec, "DM", decode = "to_decode", conformance = "off")
```

Three policies: fix the source value (the submission-grade answer), or
`no_match = "keep"` to carry the value through, or `no_match = "na"`.
Values matching only after whitespace/case normalization map with a
warning — clean them before submission, because `check_spec()` compares
exactly.

## "variables defines N variables more than once" (`vport_error_spec`)

The spec defines the same `(dataset, variable)` twice — every downstream
decision (which label? which type?) would be ambiguous, so construction
refuses. From `read_spec()` the message points at the source — sheet and
Excel row numbers for a P21 workbook:

```
! The spec defines 1 variable more than once.
x Sheet 'Variables' rows 276 and 280 all define ADCM.STUDYID.
i Fix the source, or keep the first definition of each with `on_duplicate = "first"`.
```

`read_spec(path, on_duplicate = "first")` unblocks you immediately;
`read_spec(path, datasets = "ADSL")` sidesteps a problem confined to
another domain entirely.

## "Cannot write column as a SAS date numeric" (`vport_error_codec`)

A character date column under `targetDataType = "integer"`: partial ISO
values (`"1951"`) cannot become SAS numerics without inventing data, so
the write aborts. Drop the spec's `targetDataType` to write the column
as ISO 8601 text (the SDTM `--DTC` convention — partials survive
verbatim), or complete the values. The full storage model is in
[ISO dates, --DTC, and partial dates](dtc-and-partial-dates.html).

## "Variable name is not valid for xpt v5" (`vport_error_codec`)

XPORT v5 (the FDA standard) allows 1–8 ASCII letters, digits, or
underscores, not starting with a digit. Rename the variable, or — for
non-submission exchange — `write_xpt(..., version = 8)` allows 32
characters. `check_spec()` flags these names before any write
(`variable_name`, `dataset_name`).

## "This XPORT file has more than one member" (`vport_error_codec`)

A transport file can hold several datasets; vport never guesses which
one you meant:

```{r}
#| eval: false
xpt_members("ae.xpt")            # list what the file holds
read_xpt("ae.xpt", member = "AE") # then pick one
```

## "x carries no vport metadata" (`vport_error_input`)

`get_meta()` (or a metadata-dependent write) found no `metadata_json`
attribute — usually because a transform rebuilt the frame. Capture and
re-sync:

```{r}
adsl <- apply_spec(cdisc_adsl, spec, "ADSL", conformance = "off")
meta <- get_meta(adsl)
elderly <- adsl[adsl$AGE > 80, ] # base subsetting drops attributes
elderly <- sync_meta(elderly, meta) # reconciled: columns, records, keys
get_meta(elderly)@dataset$records
```

## Catching vport conditions programmatically

Every condition class is stable API. The kinds: `input`, `spec`, `type`,
`codelist`, `codec`, `validation`, `conformance` (errors), and the
matching `vport_warning_*` / `vport_message_*` families.

```{r}
tryCatch(
  read_xpt("does-not-exist.xpt"),
  vport_error_input = function(e) "handled: bad input path"
)
```
