---
title: "How artoo compares"
format: html
---

```{r}
#| include: false
library(artoo)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

artoo is built around one idea: take a **specification** and a **data frame**,
conform the data to it, and write it out **losslessly** — to whichever
CDISC-relevant format you need, from a single package.

The tools below overlap with parts of that path. They are mature, widely used,
and each excellent at what it focuses on. This page is **orientation, not a
benchmark** — a way to see where artoo fits next to them, so you can pick the
right tool for the job in front of you.

## What each tool is built around

| Tool | Built around | A natural fit when… |
|---|---|---|
| **artoo** | the whole spec → conform → validate → write loop, in one package, lossless across formats | you want one tool from spec to file, and you need more than one output format from the same spec |
| **metacore + metatools + xportr** | a composable pharmaverse pipeline — a spec object (metacore), spec-driven helpers (metatools), and metadata-application + XPT writing (xportr) | you're already working in an admiral/pharmaverse pipeline and want building blocks that compose with it |
| **haven** | fast, faithful reading and writing of SAS/Stata/SPSS files | you simply need to read or write an XPT/SAS file, without a spec layer |
| **datasetjson** | the Dataset-JSON format specifically | Dataset-JSON is your format and you want a focused implementation of it |

These overlap rather than compete: artoo and the metacore stack both read the
same CDISC specs; artoo and haven both read and write XPT.

## What a conform-and-write looks like in artoo

The example below runs as-is on the bundled demo data — a spec plus a data frame
to a conformed, written XPT:

```{r}
adsl <- apply_spec(cdisc_adsl, adam_spec, "ADSL")
write_xpt(adsl, tempfile(fileext = ".xpt"))
```

`apply_spec()` runs the whole metadata pass in one fixed, documented order, and
is **lossless by default**: if applying the spec would lose data (for example,
coercing a fractional value to `integer`), it pauses and names the option to
continue rather than truncating silently — see
[common errors](errors.html).

The same idea in the metacore/xportr style is a short pipeline of focused steps —
both approaches are valid; they just package the work differently:

```{r}
#| eval: false
# illustrative — the pharmaverse building-block style
mc <- metacore::spec_to_metacore("adam-spec.xlsx") |>
  metacore::select_dataset("ADSL")

adsl <- raw_adsl |>
  metatools::drop_unspec_vars(mc) |>
  xportr::xportr_type(mc) |>
  xportr::xportr_label(mc) |>
  xportr::xportr_order(mc) |>
  xportr::xportr_write("adsl.xpt")
```

## Choosing by fit

- **Reach for haven** when the job is reading or writing a single file and a spec
  layer would be overhead. artoo's own `read_xpt()`/`write_xpt()` cover the same
  ground once you're already in artoo.
- **Reach for datasetjson** when Dataset-JSON is the whole task.
- **Reach for the metacore + metatools + xportr stack** when you're inside a
  pharmaverse/admiral pipeline and want components that slot into it, with the
  large body of examples and team familiarity that comes with them.
- **Reach for artoo** when you'd like one package to carry the spec → conform →
  validate → write path, and especially when you need several output formats
  (XPT, Dataset-JSON, NDJSON, Parquet) from the *same* spec.

## Using them together

Adopting artoo isn't all-or-nothing. Because it reads the same CDISC specs and the
same file formats, it sits comfortably alongside what you already use:

- Keep your admiral derivations; use artoo only for the conform-and-write step.
- Or use artoo purely for I/O (`read_xpt`/`write_xpt`/`read_json`) in an existing
  pipeline, and add `apply_spec()` later if it helps.

The goal isn't to replace good tools — it's to give you one lossless path when
that's what you want.

## Where to next

- [Get started](get-started.html) — the round-trip from the top.
- [Validation & qualification](validation.html) — the evidence a regulated
  pipeline needs.
