---
title: "Conform & validate"
format: html
---

```{r}
#| include: false
library(artoo)
```

`apply_spec()` is the heart of artoo: it conforms a raw frame to a spec
&mdash; coerce, order, sort, stamp &mdash; and never silently damages data.
Validation is the same surface read the other way: `check_spec()` and
`check_study()` return every finding at once, and every abort artoo raises
is a classed condition that names its fix. This article covers both.

## 1. Conform with `apply_spec()`

`apply_spec()` runs the same five steps, in the same documented order
&mdash; scaffold the spec's columns, coerce each to its CDISC data type,
order to the spec, sort by the dataset keys, stamp the metadata &mdash; and
returns a frame ready to write. A variable the spec declares but the data
lacks is reported, never fabricated as an empty column:

```{r}
adsl <- apply_spec(cdisc_adsl, adam_spec, "ADSL")
```

Two arguments cover the common cases. `extra = "drop"` trims columns the
spec does not mention &mdash; announced, and recorded by the `extra_variable`
finding, so the drop is never silent:

```{r}
adsl_lean <- apply_spec(cdisc_adsl, adam_spec, "ADSL", extra = "drop")
ncol(adsl_lean) <= ncol(cdisc_adsl)
```

`na_position` controls where missing key values sort. The default
`"first"` matches SAS `PROC SORT` (and FDA submission datasets); set
`"last"` only when your comparison target is R's `order()`:

```{r}
nrow(apply_spec(cdisc_adsl, adam_spec, "ADSL", na_position = "last"))
```

## 2. Lossless or loud

artoo's core rule is that no operation silently damages data. A coercion
that would truncate fractions or overflow R's 32-bit integer range aborts
with `artoo_error_type` **before** touching a value &mdash; and this gate is
independent of `conformance`, so turning conformance off does not bypass
it:

```{r}
#| error: true
#| message: false
vars <- spec_variables(adam_spec)
vars$data_type[vars$variable == "AGE"] <- "integer"
strict <- artoo_spec(
  adam_spec@datasets, vars,
  codelists = adam_spec@codelists,
  study = spec_study(adam_spec)
)
raw <- cdisc_adsl
raw$AGE[1] <- raw$AGE[1] + 0.5
apply_spec(raw, strict, "ADSL", conformance = "off")
```

You have two honest one-line exits: keep the wider source type with
`apply_spec(on_coercion_loss = "keep")` (the value is preserved and the
mismatch is left as an `integer_fraction` finding), or retype the spec with
`set_type()`.

## 3. See every finding, without halting

`apply_spec()` stops at the first thing that would lose data; to *list*
every issue instead, ask. `check_spec()` returns all of them as a tidy
frame &mdash; one row per finding, no abort:

```{r}
findings <- check_spec(cdisc_adsl, adam_spec, "ADSL")
findings
```

The same report rides along on a conformed result under the default
`conformance = "warn"`; read it back with `conformance()`:

```{r}
nrow(conformance(adsl))
```

Scaling to a whole study is one call: `check_study()` takes a named list of
datasets and returns the same shape, so "is my study submittable?" has a
single answer.

```{r}
nrow(check_study(adam_spec, list(ADSL = cdisc_adsl)))
```

## 4. Scope the checks

`artoo_checks()` toggles each conformance dimension on or off; pass the
result to `check_spec()` / `check_study()` to narrow a run to what you care
about (here, everything except the type-mismatch note):

```{r}
ck <- artoo_checks(type_mismatch = FALSE)
nrow(check_spec(cdisc_adsl, adam_spec, "ADSL", checks = ck))
```

## 5. The error family

Every abort artoo raises carries a class of the form `artoo_error_<kind>`
(plus `artoo_error` and `artoo_condition`, so one
`tryCatch(artoo_error = )` catches them all), and the data-protection
conditions carry their evidence as data, so a qualification harness can
assert on `cnd$variables` / `cnd$findings` rather than match message text.
The kinds, each triggered live:

`artoo_error_input` &mdash; the call is malformed (here, an unknown dataset):

```{r}
#| error: true
#| message: false
apply_spec(cdisc_dm, sdtm_spec, "NOPE")
```

`artoo_error_conformance` &mdash; `apply_spec(conformance = "abort")` met
error-severity findings:

```{r}
#| error: true
#| message: false
dm <- cdisc_dm
dm$SEX[1] <- "X9"
apply_spec(dm, sdtm_spec, "DM", conformance = "abort")
```

`artoo_error_codelist` &mdash; `decode_column()` met a value outside the
codelist's terms (pass `no_match = "keep"` / `"na"` to carry it through):

```{r}
#| error: true
#| message: false
decode_column(dm, sdtm_spec, "DM", from = "SEX", to = "SEXDECD")
```

`artoo_error_codec` &mdash; the bytes cannot travel (an invalid-UTF-8 value
for Dataset-JSON); re-read with the true `encoding=`, or set an
`on_invalid` policy on the writer:

```{r}
#| error: true
#| message: false
clean <- apply_spec(cdisc_dm, sdtm_spec, "DM", conformance = "off")
clean$USUBJID[1] <- rawToChar(as.raw(c(0x63, 0xE9)))
write_json(clean, tempfile(fileext = ".json"))
```

(The remaining kind, `artoo_error_spec`, fires at spec construction when a
slot references an unknown dataset or codelist, mixes standards, or
duplicates a definition.)

## Where to next

- [Specifications](specs.html) &mdash; build, inspect, and repair the spec
  this verb consumes.
- [Formats & lossless conversion](convert.html) &mdash; write the conformed
  frame to any format, and the qualification evidence behind "lossless".
- [Recipes](recipes.html) &mdash; conform inside an end-to-end ADaM and SDTM
  build.
- [Get started](artoo.html) &mdash; the round-trip from the top.
