---
title: "Formats & lossless conversion"
format: html
---

```{r}
#| include: false
library(artoo)
```

artoo's conversion story is one sentence: every codec reads and writes the
same canonical metadata, so any format converts to any other without loss.
This article shows the round trips, what "lossless" means concretely, how
the encoding policy keeps a single bad byte from deciding which formats your
dataset can travel in, and the evidence a regulated pipeline can point to.

## 1. One metadata model, four carriers

A conformed frame carries its `artoo_meta`; each writer embeds it in the
format's own idiom (XPORT NAMESTR records, the Dataset-JSON itemGroup, a
Parquet key-value sidecar, rds attributes):

```{r}
dm <- apply_spec(cdisc_dm, sdtm_spec, "DM", conformance = "off")
xpt <- tempfile(fileext = ".xpt")
json <- tempfile(fileext = ".json")
parquet <- tempfile(fileext = ".parquet")
dm |>
  write_xpt(xpt) |>
  write_json(json) |>
  write_parquet(parquet)
```

Conversion is read one, write the other — and the metadata survives the
hop:

```{r}
from_xpt <- read_xpt(xpt)
identical(get_meta(from_xpt)@columns, get_meta(read_json(json))@columns)
```

One caveat is structural, not artoo's: XPORT v5 cannot carry
`keySequence`, codelist references, or origin (its NAMESTR record has no
field for them), so an `.xpt`-sourced frame shows a blank Key pane by
design. Route through Dataset-JSON or Parquet when those must survive.

## 2. Encodings: recorded, inherited, never silently transformed

artoo text is always UTF-8 in memory; encodings matter only at the file
boundary. Each reader records the source encoding in the metadata, and
`write_*(encoding = NULL)` inherits it, so a round trip is byte-faithful.
The name rosetta lives in `artoo_encodings()`:

```{r}
enc <- artoo_encodings()
enc[enc$sas == "WLATIN1", ]
```

## 3. `on_invalid`: one policy, all four writers

A value that cannot be represented in the target — an unencodable
character for xpt's target charset, or bytes that are not valid UTF-8 for
Dataset-JSON / NDJSON / Parquet — hits the same policy everywhere:
`"error"` (default), `"replace"`, `"ignore"`, or `"translit"` (fold smart
punctuation to its exact ASCII form; see the
[WLATIN1 → UTF-8 migration article](migrate-encoding.html)). The default
names the offenders hex-escaped:

```{r}
#| error: true
bad <- dm
bad$USUBJID[1] <- rawToChar(as.raw(c(0x63, 0xE9))) # a stray latin1 byte
write_json(bad, tempfile(fileext = ".json"))
```

`"replace"` substitutes `?` and warns, so the write completes and the
file is valid:

```{r}
#| warning: true
fixed <- tempfile(fileext = ".json")
write_json(bad, fixed, on_invalid = "replace")
read_json(fixed)$USUBJID[1]
```

The right fix is upstream — re-read the source with its true `encoding=`
so the bytes arrive as the characters they were — but the policy means a
mis-declared byte in one record no longer makes a dataset "submittable as
XPT but not as Dataset-JSON", or vice versa.

## 4. Qualification: the evidence behind "lossless"

The first question a pharma organization asks of any package in a
submission pipeline is "can we qualify this?". None of the below is a
regulatory claim &mdash; qualification is your process &mdash; but every
property is machine-checkable in your own environment.

**Lossless or loud.** Read-write-read returns the same values and the same
metadata; a lossy coercion or an unencodable byte aborts rather than
damaging data; even the explicit `extra = "drop"` trim announces itself and
leaves a finding. The data-protection conditions carry their evidence as
data, so a harness asserts on it programmatically:

```{r}
vars <- spec_variables(adam_spec)
vars$data_type[vars$variable == "AGE"] <- "integer"
strict <- artoo_spec(
  adam_spec@datasets, vars,
  codelists = adam_spec@codelists,
  study = spec_study(adam_spec)
)
raw <- cdisc_adsl
raw$AGE[1] <- raw$AGE[1] + 0.5
tryCatch(
  apply_spec(raw, strict, "ADSL", conformance = "off"),
  artoo_error_type = function(cnd) cnd$variables
)
```

**The development gates**, enforced on every change and in CI: `R CMD
check` at 0 errors / 0 warnings / 0 notes; a test suite in the thousands of
assertions, including byte-level golden files for the codecs, a
cross-format round-trip matrix, and fuzzed-input tests that assert every
failure is a classed artoo condition; line coverage of at least 95% on
every file in `R/`; and reproducible demo data, every bundled object
rebuilt by script from public, checksum-pinned sources.

**Standards the behavior is pinned to**, cited in the docs rather than
re-invented: CDISC Dataset-JSON v1.1 (type vocabulary, UTF-8), the FDA
Study Data Technical Conformance Guide (XPORT expectations, ASCII gate),
SAS XPORT v5/v8 (TS-140/TS-340), RFC 8259, IANA character set names, and
Unicode NFC (UAX #15).

## Where to next

- [Specifications](specs.html) &mdash; the spec whose metadata every codec
  carries.
- [Conform & validate](conform.html) &mdash; produce the conformed frame
  these writers persist.
- [Recipes](recipes.html) &mdash; conversion inside an end-to-end ADaM and
  SDTM build.
- [Get started](artoo.html) &mdash; the round-trip from the top.
