---
title: "Any-to-any conversion"
format: html
---

```{r}
#| include: false
library(artoo)
```

artoo's conversion story is one sentence: every codec reads and writes the
same canonical metadata, so any format converts to any other without loss.
This article shows the round trips, what "lossless" means concretely, and
how the encoding policy keeps a single bad byte from deciding which
formats your dataset can travel in.

## 1. One metadata model, four carriers

A conformed frame carries its `artoo_meta`; each writer embeds it in the
format's own idiom (XPORT NAMESTR records, the Dataset-JSON itemGroup, a
Parquet key-value sidecar, rds attributes):

```{r}
dm <- apply_spec(cdisc_dm, sdtm_spec, "DM", conformance = "off")
xpt <- tempfile(fileext = ".xpt")
json <- tempfile(fileext = ".json")
parquet <- tempfile(fileext = ".parquet")
dm |>
  write_xpt(xpt) |>
  write_json(json) |>
  write_parquet(parquet)
```

Conversion is read one, write the other — and the metadata survives the
hop:

```{r}
from_xpt <- read_xpt(xpt)
identical(get_meta(from_xpt)@columns, get_meta(read_json(json))@columns)
```

One caveat is structural, not artoo's: XPORT v5 cannot carry
`keySequence`, codelist references, or origin (its NAMESTR record has no
field for them), so an `.xpt`-sourced frame shows a blank Key pane by
design. Route through Dataset-JSON or Parquet when those must survive.

## 2. Encodings: recorded, inherited, never silently transformed

artoo text is always UTF-8 in memory; encodings matter only at the file
boundary. Each reader records the source encoding in the metadata, and
`write_*(encoding = NULL)` inherits it, so a round trip is byte-faithful.
The name rosetta lives in `artoo_encodings()`:

```{r}
enc <- artoo_encodings()
enc[enc$sas == "WLATIN1", ]
```

## 3. `on_invalid`: one policy, all four writers

A value that cannot be represented in the target — an unencodable
character for xpt's target charset, or bytes that are not valid UTF-8 for
Dataset-JSON / NDJSON / Parquet — hits the same three-way policy
everywhere: `"error"` (default), `"replace"`, or `"ignore"`. The default
names the offenders hex-escaped:

```{r}
#| error: true
bad <- dm
bad$USUBJID[1] <- rawToChar(as.raw(c(0x63, 0xE9))) # a stray latin1 byte
write_json(bad, tempfile(fileext = ".json"))
```

`"replace"` substitutes `?` and warns, so the write completes and the
file is valid:

```{r}
#| warning: true
fixed <- tempfile(fileext = ".json")
write_json(bad, fixed, on_invalid = "replace")
read_json(fixed)$USUBJID[1]
```

The right fix is upstream — re-read the source with its true `encoding=`
so the bytes arrive as the characters they were — but the policy means a
mis-declared byte in one record no longer makes a dataset "submittable as
XPT but not as Dataset-JSON", or vice versa.

## Where to next

- [An end-to-end ADaM build](adam-build.html) — the derivation loop.
- [Validation & qualification](validation.html) — what "lossless" buys
  in a regulated setting.
