---
title: "Dates, times, and --DTC"
format: html
---

```{r}
#| include: false
library(artoo)
```

Clinical dates travel in two shapes — ISO 8601 text in SDTM (`--DTC`) and
numeric SAS dates in ADaM (`ADT`, `TRTSDT`) — and most tooling pain comes
from conflating them. artoo keeps the distinction explicit through three
metadata fields: `dataType`, `targetDataType`, and `displayFormat`. This
article shows how each shape is represented, realized in R, and
round-tripped. **Scope note:** this is about *carriage*. Partial-date
imputation is an analysis decision your SAP owns; artoo never imputes.

## 1. ISO text stays text (`--DTC`)

A spec variable typed `date`/`datetime` with **no** `targetDataType` is,
by the CDISC storage rule, ISO 8601 text — the SDTM `--DTC` shape.
`apply_spec()` leaves it as character, partial values included, because
`"2024-03"` is a legal ISO date and truncating or padding it would be
imputation by stealth:

```{r}
dm <- apply_spec(cdisc_dm, sdtm_spec, "DM", conformance = "off")
class(dm$BRTHDTC)
```

## 2. Typed dates realize to R classes

A variable whose `dataType` is `date`, `datetime`, or `time` realizes to
the matching R class — `Date`, `POSIXct` (UTC), `hms::hms` — in memory,
both through `apply_spec()` and on every read:

```{r}
adsl <- apply_spec(cdisc_adsl, adam_spec, "ADSL", conformance = "off")
class(adsl$TRTSDT)
```

How the value travels *in the file* is the spec's call: with no
`targetDataType` the exchange form is the ISO 8601 string; declaring
`targetDataType = "integer"` opts into SAS-epoch numbers (the ADaM
numeric-date convention). Either way, the SAS epoch (1960-01-01) versus
R epoch (1970-01-01) conversion — 3653 days — happens inside the codecs;
you never shift values yourself.

## 3. Times are `hms`

SAS `TIME` values import as `hms::hms` (seconds since midnight), which
prints as a clock time but is numeric underneath, so elapsed times past
24 hours, negative values, and fractional seconds all survive:

```{r}
hms::as_hms(30615)
```

## 4. Round trips preserve the exchange form

On write, each codec emits the exchange form the metadata dictates: ISO
strings for text-shaped temporals, SAS-epoch numbers when
`targetDataType` says numeric, with `displayFormat` (`DATE9.`,
`DATETIME20.`) riding along for SAS consumers:

```{r}
p <- tempfile(fileext = ".json")
write_json(adsl, p)
meta <- get_meta(read_json(p))
meta@columns$TRTSDT[c("dataType", "displayFormat")]
```

```{r}
identical(read_json(p)$TRTSDT, adsl$TRTSDT)
```

## Where to next

- [An end-to-end ADaM build](adam-build.html) — the full loop these
  columns live in.
- [Common errors](errors.html) — including what happens when a temporal
  value cannot be realized.
