---
title: "ISO dates, --DTC, and partial dates"
format: html
---

```{r}
#| include: false
library(artoo)
```

Clinical dates come in two storage shapes, and mixing them up corrupts a
submission. artoo encodes the distinction the way CDISC does — in the
metadata — so every codec agrees and partial dates survive byte for byte.

## The rule: `targetDataType` decides

A `dataType` of `date`, `datetime`, or `time` says what a variable
*means*. How it is *stored* follows `targetDataType`:

| Spec says | Data column is | Stored as |
|---|---|---|
| `date` (no targetDataType) | character | **ISO 8601 text** — the SDTM `--DTC` convention; partials first-class |
| `date` (no targetDataType) | `Date` / numeric | SAS-epoch numeric; `targetDataType = "integer"` is recorded at stamp time so every format agrees |
| `date` + `targetDataType = "integer"` | `Date` / numeric | SAS-epoch numeric with a SAS display format — the ADaM numeric-date convention |
| `date` + `targetDataType = "integer"` | character | **error** — a partial date can never silently become a SAS number |

This mirrors Dataset-JSON v1.1 exactly: no `targetDataType` means the
values are exchanged as ISO 8601 strings; `targetDataType: "integer"`
means SAS-epoch numbers.

## SDTM `--DTC`: character ISO text, partials included

`BRTHDTC` typed `date` with no `targetDataType` holds ISO text. Partial
dates — a known year, a known year-month — are real data, not errors:

```{r}
spec <- artoo_spec(
  data.frame(dataset = "DM", label = "Demographics"),
  data.frame(
    dataset  = "DM",
    variable  = c("USUBJID", "BRTHDTC"),
    label     = c("Subject ID", "Birth Date/Time"),
    data_type = c("string", "date")
  )
)

raw <- data.frame(
  USUBJID = c("01-001", "01-002", "01-003", "01-004"),
  BRTHDTC = c("1951-12-03", "1951-12", "1951", NA)
)

dm <- apply_spec(raw, spec, "DM")
str(dm$BRTHDTC)
```

The column stays character — artoo never silently promotes ISO text to
`Date` (a promotion would have to destroy `"1951-12"`). And it writes to
every format, XPT included, with the values intact:

```{r}
p <- tempfile(fileext = ".xpt")
write_xpt(dm, p)
read_xpt(p)$BRTHDTC
```

In the `.xpt` file `BRTHDTC` is a character variable, which is exactly
what an SDTM transport file is supposed to contain.

## ADaM numeric dates: `Date` in, SAS number out

The bundled ADSL has R `Date` columns. Conforming records
`targetDataType = "integer"` on them, so every codec stores SAS-epoch
numbers and every reader realizes the same `Date` class back:

```{r}
spec_pilot <- artoo_spec(
  cdisc_datasets, cdisc_variables,
  codelists = cdisc_codelists
)
adsl <- apply_spec(cdisc_adsl, spec_pilot, "ADSL")
get_meta(adsl)@columns$TRTSDT[c("dataType", "targetDataType")]
```

```{r}
pj <- tempfile(fileext = ".json")
write_json(adsl, pj)
identical(read_json(pj)$TRTSDT, adsl$TRTSDT)
```

## Validation: `iso8601_format`

A character temporal column is validated as ISO 8601. Complete values,
right-truncated partials, and the SDTMIG hyphen placeholders
(`"2003---15"`: day known, month unknown) all pass; a SAS-style date or
an impossible calendar date is an error-severity finding:

```{r}
bad <- raw
bad$BRTHDTC[1] <- "03DEC1951"
bad_dm <- apply_spec(bad, spec, "DM", conformance = "off")
f <- check_spec(bad_dm, spec, "DM")
f[f$check == "iso8601_format", c("severity", "variable", "message")]
```

## When the spec demands a number it cannot have

`targetDataType = "integer"` under a column that still holds partial
text is the one combination with no faithful answer — so it fails loud
at write time, never silently:

```{r}
#| error: true
spec_num <- artoo_spec(
  data.frame(dataset = "DM", label = "Demographics"),
  data.frame(
    dataset = "DM", variable = "BRTHDTC", label = "Birth Date/Time",
    data_type = "date", target_data_type = "integer"
  )
)
dm_num <- apply_spec(raw["BRTHDTC"], spec_num, "DM", conformance = "off")
write_xpt(dm_num, tempfile(fileext = ".xpt"))
```

The fix is in the message: drop the `targetDataType` to keep ISO text,
or complete the values.
