---
title: "Migrating from metacore, metatools, and xportr"
format: html
---

```{r}
#| include: false
library(vport)
```

If you build submission datasets with the pharmaverse metadata stack —
`metacore` to hold the spec, `metatools` to derive from it, `xportr` to
stamp and write — vport replaces all three with one object and one
pipeline. This article is the Rosetta stone: every call you use today,
and the vport equivalent.

## The Rosetta table

| You write today | vport equivalent | Notes |
|---|---|---|
| `metacore::spec_to_metacore(path)` | `read_spec(path)` | Reads P21 Excel, Define-XML 2.x, and vport's native JSON through one validated constructor. |
| `metacore::select_dataset(mc, "ADSL")` | `read_spec(path, datasets = "ADSL")` | Scoped at read time, so a broken sheet elsewhere in the workbook cannot block you. |
| `metatools::build_from_derived(mc, ds_list)` | `apply_spec(raw, spec, "ADSL")` (scaffold + drop steps) | One call runs the whole conform pipeline; `steps =` selects a subset. |
| `metatools::create_var_from_codelist(df, mc, RACE, RACEN)` | `decode_column(df, spec, "ADSL", from = "RACE", to = "RACEN", direction = "to_code")` | The destination's codelist wins; a spec-declared destination is coerced to its dataType and labelled. |
| `metatools::check_variables(df, mc)` | `check_spec(df, spec, "ADSL")` | One findings frame covers variables, types, lengths, codelists, keys, ISO 8601 dates. |
| `metatools::check_ct_data(df, mc)` | `check_spec()` (codelist dimensions) or `validate_spec(spec, data = df)` | Extensible codelists downgrade non-members to notes automatically. |
| `metatools::order_cols(df, mc)` + `sort_by_key(df, mc)` | `apply_spec(..., steps = c("order", "sort"))` | Or let the full pipeline do both. |
| `xportr::xportr_type/length/label/format/order(...)` | `apply_spec(raw, spec, "ADSL")` | The metadata stamp carries label, length, format, and order in one `vport_meta`. |
| `xportr::xportr_write(df, path)` | `write_xpt(adsl, "adsl.xpt")` | Plus `write_json()`, `write_parquet()`, `write_rds()` from the same object. |

## A migration, end to end

The bundled CDISC pilot data stands in for your study. The whole legacy
chain collapses to three pipeable calls:

```{r}
spec <- vport_spec(
  cdisc_datasets,
  cdisc_variables,
  codelists = cdisc_codelists
)

adsl <- cdisc_adsl |>
  apply_spec(spec, "ADSL") |>
  write_xpt(tempfile(fileext = ".xpt")) |>
  write_json(tempfile(fileext = ".json"))
```

`apply_spec()` scaffolds missing spec variables, drops undeclared
columns, coerces every column to its CDISC dataType, orders, sorts by the
key sequence, and stamps the metadata. The `write_*()` calls return their
input invisibly, so a pipeline can fan one conformed frame out to every
deliverable format.

## The `create_var_from_codelist()` pattern

Deriving a coded numeric from its decode — `RACEN` from `RACE` — is
`decode_column()` with `direction = "to_code"`. Here, a `SEXN` variable
declared as `integer` with its own numeric codelist:

```{r}
vars <- rbind(
  cdisc_variables,
  data.frame(
    dataset = "DM", variable = "SEXN", label = "Sex (N)",
    data_type = "integer", length = 8L, order = NA_integer_,
    codelist_id = "SEXN"
  )
)
cls <- rbind(
  cdisc_codelists,
  data.frame(
    codelist_id = "SEXN", term = c("1", "2"),
    decode = c("F", "M"), order = 1:2
  )
)
spec_n <- vport_spec(cdisc_datasets, vars, codelists = cls)

dm <- decode_column(
  cdisc_dm, spec_n, "DM",
  from = "SEX", to = "SEXN", direction = "to_code"
)
str(dm$SEXN)
attr(dm$SEXN, "label")
```

Because `SEXN` is declared in the spec, the result arrives as an
`integer` with its label attached — no follow-up `xportr_type()` /
`xportr_label()` pass.

## Matching your legacy pipeline's output exactly

During a migration you often need byte-comparable output before you
switch the defaults on. `profile = "xportr"` reproduces the legacy
pipeline's shape — drop, order, sort, stamp; **no** scaffolding, **no**
type coercion, **no** decode:

```{r}
legacy_shape <- apply_spec(
  cdisc_dm, spec, "DM",
  profile = "xportr",
  conformance = "off"
)
```

Once the outputs reconcile, drop the profile and let the full pipeline
(including the coercion and the lossy-coercion guard) take over.

## What you gain on the other side

- **One metadata object.** The `vport_meta` stamped by `apply_spec()` is
  exactly what every codec writes and reads back — there is no second
  metadata pass and no drift between formats.
- **Findings, not console noise.** `conformance(adsl)` returns the full
  findings frame with a sectioned print; severity is data you can gate a
  pipeline on.
- **Loud data-integrity failures.** A spec that would truncate
  `HEIGHTBL` from 162.6 to 162 aborts before any value is damaged
  (`on_lossy = "error"`, the default) instead of warning after the fact.
- **Partial dates are first-class.** A character `--DTC` column writes to
  XPT as ISO text, `"1951-12"` included — see
  [ISO dates, --DTC, and partial dates](dtc-and-partial-dates.html).
