---
title: "Migrating clinical data from WLATIN1 to UTF-8"
format: html
---

```{r}
#| include: false
library(artoo)
```

Most legacy clinical data was produced by SAS sessions running WLATIN1
(Windows-1252); modern pipelines — Dataset-JSON, Parquet, R itself — are
UTF-8. The two agree on every ASCII character and disagree everywhere
else, so a migration has exactly three failure modes: mojibake (bytes
interpreted as the wrong charset), truncation (byte-counted lengths that
multibyte UTF-8 overflows), and unrepresentable characters (an ASCII
target that cannot carry what UTF-8 can). This article shows how artoo
handles each one.

## 1. Reading WLATIN1 data

`read_xpt()` detects the encoding: if every byte in the file validates
as UTF-8 it reads as UTF-8, otherwise it assumes Windows-1252 — the
right default for legacy clinical files. Formats whose spec pins UTF-8
(Dataset-JSON, Parquet) are read as UTF-8 and *refuse loudly* when the
bytes disagree, rather than guessing:

```{r}
#| error: true
latin1 <- function(x) {
  out <- iconv(x, "UTF-8", "windows-1252")
  Encoding(out) <- "unknown"
  out
}
foreign <- data.frame(SITEID = latin1(c("PARIS", "MÜNCHEN")))
pq <- tempfile(fileext = ".parquet")
nanoparquet::write_parquet(foreign, pq)
read_parquet(pq)
```

That refusal is the prompt to name the true source charset. `encoding=`
is a *transcode*, not a label — the bytes are converted to UTF-8 and
normalised (NFC), so what lands in R is ordinary text:

```{r}
site <- read_parquet(pq, encoding = "wlatin1")
site$SITEID
```

Any SAS or IANA spelling works — `"wlatin1"`, `"cp1252"`,
`"windows-1252"` are the same charset; so are the OEM/DOS names
(`"pcoem850"`) a very old archive might declare. The rosetta is
`artoo_encodings()`. Note that `"latin1"` (ISO-8859-1) is **not**
WLATIN1: the two differ across the 0x80–0x9F range that holds the Euro
sign and every smart quote.

## 2. Catching a mis-declared read

The classic silent corruption is reading WLATIN1 bytes *as if* they were
UTF-8 (in SAS terms: a data set whose encoding attribute lies). Those
bytes do not validate as UTF-8, and the `invalid_encoding` conformance
dimension flags them before any writer runs:

```{r}
bad <- cdisc_dm
bad$ETHNIC <- rawToChar(as.raw(c(0x4D, 0xDC, 0x4E))) # WLATIN1 bytes, raw
f <- check_spec(bad, sdtm_spec, "DM")
f[f$check == "invalid_encoding", c("variable", "severity", "message")]
```

The fix is always upstream — re-read the source with its true
`encoding=` — never a re-label.

## 3. Lengths are bytes, and UTF-8 needs more of them

A WLATIN1 character is always one byte; in UTF-8 the same character can
take up to four. A spec length calibrated on WLATIN1 data (`Ö` = 1 byte)
under-declares the UTF-8 form (`Ö` = 2 bytes). SAS handles this with the
CVP engine and a guessed multiplier; artoo measures the actual bytes,
never truncates, and tells you when the declared length had to give:

```{r}
#| warning: true
sp <- artoo_spec(
  data.frame(dataset = "DM"),
  data.frame(
    dataset = "DM", variable = "INVNAM",
    data_type = "string", length = 6L
  )
)
x <- apply_spec(
  data.frame(INVNAM = "Öztürk"), sp, "DM",
  conformance = "off"
)
xpt <- tempfile(fileext = ".xpt")
write_xpt(x, xpt)
```

The `length_overflow` check reports the same fact at conformance time,
so a migration can fix the spec (the durable answer) instead of relying
on the writer's defence.

## 4. Smart punctuation: the WLATIN1 characters ASCII can fold

Text that passed through a word processor carries typographic
punctuation — curly quotes, en/em dashes, ellipses. In WLATIN1 these
live in rows 8–9 of the code page (one byte each); in UTF-8 they are
three bytes each; in US-ASCII they do not exist at all. They are also
the one class of character with an *exact* ASCII equivalent, published
by SAS as the NLS punctuation fold:

| WLATIN1 (hex) | Unicode | Character | Description | ASCII fold |
|---|---|:---:|---|:---:|
| `82` | U+201A | ‚ | single low-9 quotation mark | `,` |
| `84` | U+201E | „ | double low-9 quotation mark | `"` |
| `85` | U+2026 | … | horizontal ellipsis | `...` |
| `8B` | U+2039 | ‹ | single left-pointing angle quote | `<` |
| `91` | U+2018 | ‘ | left single quotation mark | `'` |
| `92` | U+2019 | ’ | right single quotation mark | `'` |
| `93` | U+201C | “ | left double quotation mark | `"` |
| `94` | U+201D | ” | right double quotation mark | `"` |
| `95` | U+2022 | • | bullet | `*` |
| `96` | U+2013 | – | en dash | `-` |
| `97` | U+2014 | — | em dash | `-` |
| `9B` | U+203A | › | single right-pointing angle quote | `>` |

Every writer accepts `on_invalid = "translit"`, which applies exactly
this table and nothing else:

```{r}
#| warning: true
note <- data.frame(
  USUBJID = "01-701-1015",
  COMMENT = "Patient’s dose – “held”…"
)
ascii_xpt <- tempfile(fileext = ".xpt")
write_xpt(note, ascii_xpt, encoding = "US-ASCII", on_invalid = "translit")
read_xpt(ascii_xpt)$COMMENT
```

`"translit"` is deliberately narrower than `"replace"`: punctuation
folds because the ASCII form carries the same meaning; a character with
no equivalent — a diacritic in a name — still aborts, because
`"Öztürk"` silently becoming `"Ozturk"` (or worse, `"?zt?rk"`) is data
corruption, not conversion:

```{r}
#| error: true
name <- data.frame(INVNAM = "Öztürk")
write_xpt(name, tempfile(fileext = ".xpt"),
  encoding = "US-ASCII", on_invalid = "translit"
)
```

## 5. The migration recipe

1. **Read with the true source encoding.** Let `read_xpt()` detect, or
   pass `encoding = "wlatin1"` explicitly for formats that cannot carry
   the answer. Verify with `check_spec()` — zero `invalid_encoding`
   findings means the bytes arrived as characters.
2. **Re-measure lengths in bytes.** Run the `length_overflow` dimension
   against the UTF-8 form and update the spec where multibyte
   characters grew.
3. **Write UTF-8 by default** (Dataset-JSON, Parquet, xpt v8 all carry
   it losslessly). For a US-ASCII submission XPT, write with
   `encoding = "US-ASCII", on_invalid = "translit"` — punctuation
   folds, genuine data problems stay loud.

## References

- SAS 9.4 NLS Reference Guide, *Migrating Data from WLATIN1 to UTF-8*
  (the WLATIN1 code-page figures and the smart-punctuation table):
  <https://documentation.sas.com/doc/en/pgmmvacdc/9.4/nlsref/n15e31tqdv020en1fok7tp4l9zd5.htm>
- Bouedo, M. (2020), *The SAS Encoding Journey: A Byte at a Time*,
  SAS Global Forum paper 4561-2020:
  <https://www.sas.com/content/dam/SAS/support/en/sas-global-forum-proceedings/2020/4561-2020.pdf>
- FDA Study Data Technical Conformance Guide (the ASCII expectation for
  submission XPORT): <https://www.fda.gov/media/153632/download>
- Unicode Standard Annex #15, *Unicode Normalization Forms* (the NFC
  form artoo canonicalises to): <https://unicode.org/reports/tr15/>
