---
title: "Recipes"
format: html
---

```{r}
#| include: false
library(artoo)
```

These are the everyday programming loops a clinical programmer runs &mdash;
an ADaM analysis dataset, an SDTM domain, a codelist decode, the temporal
shapes &mdash; each rendered **live** below from the bundled demo data, so
what you see is exactly what the code produces. Every recipe ends on a
conformed result; the closing line shows how to **ship it** to any
deliverable, just by changing the extension.

## An ADaM build: ADSL

A real derivation program accumulates working columns the spec never
declares &mdash; here an age-group cut. Those are "extras"; `extra = "drop"`
trims them to exactly the spec's columns, recorded by the `extra_variable`
finding so the drop is never silent. `apply_spec()` then coerces, orders,
sorts, and stamps:

```{r}
#| message: false
adsl_raw <- cdisc_adsl
adsl_raw$AGEGR1_TMP <- cut(
  adsl_raw$AGE,
  breaks = c(-Inf, 64, 74, Inf),
  labels = c("<65", "65-74", ">=75")
)
adsl_raw$AGEGR1 <- as.character(adsl_raw$AGEGR1_TMP)

adsl <- apply_spec(adsl_raw, adam_spec, "ADSL", extra = "drop")
columns(adsl)
```

```{r}
#| eval: false
write_xpt(adsl, "adsl.xpt") # ship: or .json / .parquet / .rds
```

## An SDTM build: DM

The pipeline is identical &mdash; only the spec and the data change. Assemble
the domain (with a QC temporary), conform, then read the result back as a
one-line inventory with `members()`:

```{r}
#| message: false
dm_raw <- cdisc_dm
dm_raw$AGE_CHECK <- dm_raw$AGE >= 18

dm <- apply_spec(dm_raw, sdtm_spec, "DM", extra = "drop", conformance = "off")

json <- tempfile(fileext = ".json")
write_json(dm, json)
members(json)
```

```{r}
#| eval: false
write_xpt(dm, "dm.xpt") # ship: or .json / .parquet / .rds
```

## Codelists: decode in either direction

`decode_column()` reads the codelist a variable is bound to in the spec and
maps in either direction &mdash; here, deriving the numeric `RACEN` code from
the `RACE` decode (`direction = "to_code"`):

```{r}
#| message: false
adsl <- apply_spec(cdisc_adsl, adam_spec, "ADSL", conformance = "off")
coded <- decode_column(
  adsl, adam_spec, "ADSL",
  from = "RACE", to = "RACEN", direction = "to_code"
)
unique(coded[, c("RACE", "RACEN")])
```

A value outside the codelist's terms aborts (`artoo_error_codelist`) unless
you pass `no_match = "keep"` / `"na"`.

## Dates, times, and `--DTC`

Clinical dates travel in two shapes, and artoo keeps the distinction
explicit through `dataType`, `targetDataType`, and `displayFormat`. **Scope
note:** this is about *carriage* &mdash; partial-date imputation is an
analysis decision your SAP owns; artoo never imputes.

An SDTM `--DTC` variable typed `date` is ISO 8601 text by the CDISC storage
rule, so it stays character through the pipeline, partial values included
(`"2024-03"` is a legal ISO date, and padding it would be imputation by
stealth):

```{r}
#| message: false
dm <- apply_spec(cdisc_dm, sdtm_spec, "DM", conformance = "off")
class(dm$RFSTDTC)
```

A variable typed `date` with no `targetDataType` realizes to an R `Date` in
memory:

```{r}
#| message: false
adsl <- apply_spec(cdisc_adsl, adam_spec, "ADSL", conformance = "off")
class(adsl$DISONSDT)
```

The ADaM numeric-date convention is the other shape: a variable carried as
an `integer` SAS-epoch day count, with a `displayFormat` (`date9.`) telling
SAS how to render it. `TRTSDT` is one, and the format rides in the metadata:

```{r}
class(adsl$TRTSDT)

p <- tempfile(fileext = ".json")
write_json(adsl, p)
get_meta(read_json(p))@columns$TRTSDT[c("dataType", "displayFormat")]
```

SAS `TIME` values import as `hms` (seconds since midnight), so values past
24 hours, negative values, and fractional seconds all survive:

```{r}
hms::as_hms(30615)
```

Either date shape round-trips byte-faithfully &mdash; the SAS-epoch
(1960-01-01) versus R-epoch conversion happens inside the codecs, never in
your code:

```{r}
identical(read_json(p)$TRTSDT, adsl$TRTSDT)
```

## Where to next

- [Specifications](specs.html) &mdash; the spec each recipe starts from.
- [Conform & validate](conform.html) &mdash; `apply_spec()` and the findings
  in depth.
- [Formats & lossless conversion](convert.html) &mdash; the round trips these
  ship lines rely on.
- [Get started](artoo.html) &mdash; the round-trip from the top.
