---
title: "An end-to-end SDTM build"
format: html
---

```{r}
#| include: false
library(artoo)
```

This article walks the SDTM tabulation loop with artoo doing the carriage:
load the spec, assemble the domain, check controlled terminology, conform,
and fan the result out to every deliverable format. Everything runs on the
bundled demo data (`cdisc_dm`, `sdtm_spec` — built reproducibly from the
official CDISC Define-XML 2.1 example and the PHUSE Test Data Factory). It
is the SDTM counterpart of the [ADaM build](adam-build.html); the pipeline
is identical, only the spec and the data change.

## 1. The spec is the contract

`sdtm_spec` describes the SDTM domains DM, VS, TS, and SUPPDM for
SDTMIG 3.1.2 — variables, CDISC data types, lengths, labels, codelists,
and keys:

```{r}
sdtm_spec
```

The accessors return plain data frames, so the spec slots into ordinary
base R work:

```{r}
vars <- spec_variables(sdtm_spec, "DM")
head(vars[, c("variable", "label", "data_type", "codelist_id")])
```

## 2. Assemble the domain, with temporaries

A tabulation program accumulates working columns the spec never declares —
a QC flag, an intermediate join. Those are "extras", and you decide their
fate at conform time:

```{r}
dm <- cdisc_dm
dm$AGE_CHECK <- dm$AGE >= 18
```

## 3. Controlled terminology: check membership

Each coded variable's row in the spec names its `codelist_id`, and
`spec_codelists()` returns the terms a value is allowed to take:

```{r}
head(spec_codelists(sdtm_spec))
```

`apply_spec()` never translates values, so submission values stay exactly
as collected; `decode_column()` is the explicit single-variable translator
when you need to move between a code and its decode.

## 4. Conform: coerce, order, sort, stamp

`apply_spec()` runs the fixed pipeline and stamps the CDISC metadata.
`extra = "drop"` trims the QC temporaries to exactly the spec's columns —
announced, and recorded by the `extra_variable` finding, so the drop is
never silent:

```{r}
conformed <- apply_spec(dm, sdtm_spec, "DM", extra = "drop")
"AGE_CHECK" %in% names(conformed)
```

A `--DTC` reference date such as `RFSTDTC` is ISO 8601 text by CDISC
definition, so it stays character through the pipeline; in the
`PROC CONTENTS`-style pane its length is the stored byte width, while a
numeric carries none:

```{r}
columns(conformed)
```

## 5. Fan out to every deliverable

The writers return their input invisibly, so one conformed domain flows to
every format in a single pipe — losslessly, because each codec carries the
same metadata. `members()` then reads any container back as a one-line
inventory:

```{r}
xpt <- tempfile(fileext = ".xpt")
json <- tempfile(fileext = ".json")
conformed |>
  write_xpt(xpt) |>
  write_json(json)
members(json)
```

## Where to next

- [An end-to-end ADaM build](adam-build.html) — the analysis-dataset loop.
- [Any-to-any conversion](convert.html) — moving between the formats.
- [Dates, times, and `--DTC`](dates.html) — how temporal values travel.
- [Common errors](errors.html) — every condition, its trigger, its fix.
