---
title: "Specifications"
format: html
---

```{r}
#| include: false
library(artoo)
```

A `artoo_spec` is artoo's single source of truth: the variables, CDISC data
types, lengths, labels, controlled-terminology codelists, and sort keys for
exactly **one** CDISC standard. Read one from the metadata you already have,
inspect it as plain data frames, fix it in R when the data disagrees, and
write it back &mdash; the spec is the contract every later step honors.

![The artoo lossless round-trip, with the spec and its fix-and-inspect verbs feeding apply_spec.](../man/figures/round-trip-hero.svg){fig-alt="A spec plus data flow into apply_spec; set_type and check_spec feed in to fix and inspect the spec before it is applied." width="100%"}

## 1. Read a spec

`read_spec()` ingests a specification from Define-XML 2.x, a Pinnacle 21
workbook, or artoo's own native JSON, and returns a `artoo_spec`. The bundled
ADaM spec also ships as a P21 workbook, so this runs as-is:

```{r}
p21 <- system.file("extdata", "adam-spec.xlsx", package = "artoo")
spec <- read_spec(p21)
spec
```

A workbook can carry several standards or duplicate roles; scope the read
when you need just one:

```r
read_spec("define.xml", datasets = "ADSL", on_duplicate = "first")
```

## 2. Inspect with the `spec_*` accessors

Each accessor returns a plain data frame (or character vector), so the spec
slots straight into ordinary base R work &mdash; filter, join, summarise. The
datasets a spec covers:

```{r}
spec_datasets(spec)
```

The variable table is the one you reach for most; here, four columns of it:

```{r}
spec_variables(spec, "ADSL")[, c("variable", "label", "data_type", "length")] |>
  head()
```

The sort keys that `apply_spec()` will order by, and the controlled
terminology a coded variable is bound to:

```{r}
spec_keys(spec, "ADSL")

head(spec_codelists(spec))
```

`spec_standard()`, `spec_study()`, `spec_methods()`, `spec_comments()`, and
`spec_documents()` expose the remaining slots the same way.

## 3. Fix it in place

When the data disagrees with the spec, fix the spec in one line &mdash; never
reach into internals. `set_type()` retypes a variable; the spec is immutable,
so it returns an updated copy:

```{r}
spec <- set_type(spec, "ADSL", AGE = "float")
v <- spec_variables(spec, "ADSL")
v$data_type[v$variable == "AGE"]
```

When a check has already found integer-vs-fraction mismatches, `repair_spec()`
applies the fix for every one of them at once, from the findings frame:

```{r}
findings <- check_spec(cdisc_adsl, spec, "ADSL")
spec <- repair_spec(spec, findings)
```

## 4. Write it back

`write_spec()` is the inverse of `read_spec()` on each format: native JSON is
fully lossless, and the P21 workbook is the interchange form. Round-trip a
corrected spec through JSON:

```{r}
out <- tempfile(fileext = ".json")
write_spec(spec, out)
identical(spec_standard(read_spec(out)), spec_standard(spec))
```

Because the two verbs are inverses, format conversion is one composition:

```r
read_spec("define.xml") |> write_spec("spec.xlsx")
```

## Where to next

- [Conform & validate](conform.html) &mdash; apply this spec to data, then
  check every finding.
- [Formats & lossless conversion](convert.html) &mdash; move a conformed
  dataset between formats without loss.
- [Recipes](recipes.html) &mdash; the spec in an end-to-end ADaM and SDTM
  build.
- [Get started](artoo.html) &mdash; the round-trip from the top.
