---
title: "Validation and qualification evidence"
format: html
---

```{r}
#| include: false
library(vport)
```

"Can I use this for a submission?" is the first question a validated
environment asks. Validation is your organisation's process — no package
can declare itself validated — but vport is engineered to make that
assessment short. This page collects the evidence and where to verify
each piece independently.

## The lossless guarantee, stated precisely

vport's core claim: **a dataset written and read back through any
supported format preserves the data values and the recorded `vport_meta`
attributes that the format can carry.** Self-describing formats
(Dataset-JSON, Parquet, rds) carry the complete metadata; SAS XPORT
carries what its NAMESTR physically stores (name, label, length, format)
— and that boundary is documented per codec rather than papered over
(see the *What XPORT can carry* section of `?write_xpt`).

The guarantee is enforced, not aspirational:

- **A round-trip test matrix** drives conformed CDISC pilot datasets
  through every format pair and asserts equality of values and metadata
  (`tests/testthat/test-roundtrip-matrix.R`).
- **Byte-golden files** pin the exact output bytes of the XPT and
  Dataset-JSON writers against checked-in references, so an accidental
  encoding change cannot land silently.
- **An external oracle** checks the SAS epoch arithmetic against known
  constants (1960-01-01 is SAS day 0) independent of the conversion code
  under test.
- **Cross-implementation validation**: CI reads vport-written XPT files
  with `pyreadstat` (a SAS-independent C implementation) and compares
  values, labels, and formats.
- **Adversarial inputs**: fuzzing and malformed-file tests assert that a
  truncated, bit-flipped, or hostile file produces a classed vport error,
  never silent corruption.

## Quality system facts

Each of these is continuously verified in public CI, not asserted here:

- `R CMD check` runs clean (0 errors, 0 warnings, 0 notes) on every
  commit, across platforms, on GitHub Actions.
- The testthat suite (thousands of assertions) gates every merge; the
  coverage badge on the README links to per-file line coverage on
  Codecov. The project's working standard is at least 95% line coverage
  per source file.
- Every conformance and validation finding is keyed to an **open rule
  catalog** shipped in the package (`spec_rules.json` in `inst/`), so the
  behavior of `check_spec()` / `validate_spec()` is enumerable and
  auditable:

```{r}
rules <- jsonlite::fromJSON(system.file("spec_rules.json", package = "vport"))
table(rules$engine, rules$severity)
```

- Output is **deterministic**: pass a frozen `created =` timestamp and
  two writes of the same data are byte-identical — the property your
  re-execution / double-programming checks need.
- Writes are **atomic** (temp file, then rename), so an interrupted write
  can never leave a half-written deliverable in place.

## What regulators actually require

- **FDA Study Data Technical Conformance Guide**: vport enforces the
  XPORT v5 constraints (8-character names, 40-byte labels, 200-byte
  character values) as conformance findings, warns on the prohibited
  byte range 160–191 in submission transport files, and defaults XPT
  writes to the v5 standard.
- **CDISC Dataset-JSON v1.1**: the metadata model *is* the Dataset-JSON
  v1.1 itemGroup vocabulary, verbatim — dataTypes, targetDataType,
  keySequence, displayFormat. Anything vport needs beyond the closed
  CDISC vocabulary rides one namespaced `_vport` block, and
  `write_json(strict = TRUE)` suppresses even that, with a warning naming
  exactly what was dropped.

## Qualifying vport in your environment

A typical risk-based qualification (R Validation Hub style) needs:

1. **Requirements coverage** — map your intended use to the exported
   functions; each help page documents behavior, error classes, and runs
   its examples under `R CMD check`.
2. **Installation qualification** — `R CMD check` the released tarball in
   your environment; the suite is self-contained (bundled demo data, no
   network, no SAS or Java runtime).
3. **Operational qualification** — run the package test suite in place:
   `testthat::test_package("vport")`. Every property above is asserted
   there.
4. **Performance qualification** — round-trip one of *your* datasets and
   compare with your existing process. `conformance()` and
   `validate_spec()` give you the findings as data frames you can file
   as evidence.

If your QA group needs anything that is not verifiable from the public
repository, [open an issue](https://github.com/vthanik/vport/issues) —
documentation gaps in the evidence trail are treated as bugs.
