---
title: "Why vport"
format: html
---

```{r}
#| include: false
library(vport)
```

There are several R packages that read or write one clinical file
format. vport exists because a submission pipeline never touches just
one format — and because the metadata (labels, CDISC types, lengths,
formats, codelists, sort keys) matters as much as the values.

## One metadata model, every format

vport carries a single CDISC-shaped metadata object — the `vport_meta`,
which is the Dataset-JSON v1.1 itemGroup vocabulary — through every
codec. Whatever format you write, the same serializer embeds the same
metadata, so **any-to-any conversion is lossless by construction**, not
by per-pair effort:

```{r}
spec <- vport_spec(cdisc_datasets, cdisc_variables, codelists = cdisc_codelists)
adsl <- apply_spec(cdisc_adsl, spec, "ADSL")

p_xpt <- tempfile(fileext = ".xpt")
p_parquet <- tempfile(fileext = ".parquet")

# xpt -> read -> parquet, metadata intact at every hop
adsl |> write_xpt(p_xpt)
read_xpt(p_xpt) |> write_parquet(p_parquet)
get_meta(read_parquet(p_parquet))@columns$STUDYID$label
```

## What the alternatives carry

|  | XPT | Dataset-JSON | Parquet | Spec-driven conform | CDISC metadata model | Compiled deps |
|---|---|---|---|---|---|---|
| **vport** | read + write, v5 + v8, multi-member | v1.1 read + write (+ NDJSON) | read + write with full metadata sidecar | yes (`apply_spec`) | Dataset-JSON v1.1, verbatim | none in vport itself |
| haven | read + write | — | — | — | label/format attributes only | C (ReadStat) |
| xport | read + write (v5) | — | — | — | minimal | — |
| datasetjson | — | read + write | — | — | Dataset-JSON | — |
| arrow / nanoparquet | — | — | read + write | — | none (drops labels) | C++ |
| metacore + metatools + xportr | write via haven | — | — | yes, across three packages | spec object (own shape) | via haven |

Two structural differences matter more than any single row:

- **The metadata survives the I/O.** A parquet file written by
  `write_parquet()` carries the complete CDISC metadata as a
  Dataset-JSON sidecar in the file's key-value metadata; plain
  arrow/nanoparquet output drops labels and formats entirely. The same
  applies to rds and, within its physical limits, XPT.
- **The spec drives the data.** `read_spec()` ingests P21 Excel,
  Define-XML 2.x, or native JSON; `apply_spec()` conforms a raw frame to
  it; `check_spec()` reports every divergence against an open rule
  catalog. In the pharmaverse that workflow spans three packages and two
  metadata representations.

## Pure R, by design

vport contains no compiled code and needs no SAS, Java, or C++ toolchain:
the XPORT codec (IBM floating point included) and the Dataset-JSON
structure are implemented in R, with `jsonlite` doing JSON tokenization
and the tiny `nanoparquet` engine handling Parquet pages. That keeps
installation trivial in locked-down validated environments — the places
clinical R actually runs.

## Honest about physical limits

Losslessness claims are only credible when the boundaries are explicit.
vport documents what each container cannot carry instead of hiding it:
XPT cannot distinguish `""` from `NA`, stores at most 40 label bytes, and
holds no CDISC metadata beyond the NAMESTR fields; v5 cannot record a row
count. Where a choice would silently damage data — a partial date under a
numeric targetDataType, an integer coercion that truncates fractions, a
value outside its codelist — vport errors loudly by default rather than
producing a plausible wrong file.

## Where to go next

- [From spec to submission dataset](from-spec-to-submission.html) —
  the core workflow.
- [Migrating from metacore, metatools, and xportr](migrate-from-pharmaverse.html)
  — the Rosetta table.
- [Validation and qualification evidence](validation-and-qualification.html)
  — for your QA group.
