---
title: "Authoring a specification workbook"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(artoo)
```

Most specifications are written in Excel, in a workbook of one sheet per
kind of metadata. artoo reads that workbook and writes the submission's
define.xml from it. This article is the reference for filling one in: what
each sheet is for, which cells artoo actually reads, and the handful of
places where a plausible-looking entry produces a document that is not what
you meant.

It is written for the person holding the workbook. If you want the R side —
inspecting a spec, repairing it, writing the define.xml — that is
[Specifications](specs.html).

## Start from a blank one

`write_template()` writes an empty workbook whose sheets and headers are
artoo's own reader, so nothing you fill in can be silently ignored:

```{r}
book <- tempfile(fileext = ".xlsx")
write_template(book)
readxl::excel_sheets(book)
```

Only **Datasets** and **Variables** are required. Every other sheet may be
left empty and artoo omits what it finds nothing in.

## Two shapes, and how to tell them apart

Workbooks come in two generations, and artoo reads both. You will recognise
which one you have from three things:

| | Older | Newer |
|---|---|---|
| The study sheet is called | `Study` | `Define` |
| Conditions live | on a `WhereClauses` sheet | in the `ValueLevel` cell |
| The label column is called | `Description` | `Label` |

If your workbook came out of a validation tool, it is whichever that tool
writes. If you are starting fresh, use the template above — it is the newer
shape — and read on; anywhere the two differ, this article says so.

## Define / Study — the study sheet

Two columns, `Attribute` and `Value`, one row per attribute.

These five are the ones to fill in:

| Attribute | Why it matters |
|---|---|
| `StudyName` | The study's short name. |
| `StudyDescription` | One line describing the study. |
| `ProtocolName` | The protocol identifier. |
| `StandardName` | `SDTM-IG`, `ADaM-IG`, `SEND-IG`. Unhyphenated spellings are understood too. |
| `StandardVersion` | `3.4`, `1.1`, and so on. |

**`StandardName` and `StandardVersion` are the two that bite.** Define-XML
2.0 refuses to be written without them, and until artoo learned to read the
pair, a workbook that named its standard perfectly clearly still could not
produce a define.xml. Fill them in.

A workbook artoo wrote from an existing define.xml carries more, because
that document had identifiers this sheet is where to keep:

```{r}
book3 <- tempfile(fileext = ".xlsx")
suppressWarnings(write_spec(adam_spec, book3))
readxl::read_excel(book3, sheet = "Define")
```

You do not have to supply any of those; artoo mints what a new document
needs. Leave them alone if they are there.

Anything else you put on this sheet is kept and written back out, so a
sponsor field of your own survives a round trip.

## Datasets — one row per dataset

`Dataset` and one of `Description` / `Label` are the minimum.

| Column | Notes |
|---|---|
| `Dataset` | The name, `DM`, `ADSL`. |
| `Description` or `Label` | The dataset label. |
| `Class` | `EVENTS`, `FINDINGS`, `BASIC DATA STRUCTURE`, ... |
| `Structure` | Prose: "One record per subject". **Required** — Define-XML has no default for it. |
| `Key Variables` | Space- or comma-separated. This is where a variable's key sequence comes from, so you do not fill that in twice. |
| `Purpose` | `Tabulation` or `Analysis`. Derived from the standard when blank. |
| `Repeating`, `Reference Data` | `Yes` / `No`. |
| `Comment` | A **Comment ID**, matching a row on the Comments sheet — not free text. |

## Variables — one row per variable per dataset

| Column | Notes |
|---|---|
| `Order` | Position within the dataset. |
| `Dataset`, `Variable` | Together they identify the row. |
| `Label` | The variable label. |
| `Data Type` | `text`, `integer`, `float`, `date`, `datetime`, `time`. |
| `Length` | Expected on text, integer and float; **leave it blank on dates and times**, where it is a conformance finding. |
| `Significant Digits` | Only with `float`. |
| `Format` | The display format. |
| `Mandatory` | `Yes` / `No`. |
| `Codelist` | A Codelist ID **or a Dictionary ID** — one column serves both. |
| `Origin` | `Collected`, `Derived`, `Assigned`, `Protocol`, `Predecessor`. |
| `Pages` | CRF page numbers. See *Pages* below. |
| `Method` | A Method ID, for a derived variable. |
| `Predecessor` | Where an ADaM variable came from, e.g. `DM.AGE`. |
| `Role` | `Identifier`, `Topic`, `Timing`, ... |
| `Comment` | A Comment ID. |

**Origin decides what else you must fill.** `Derived` wants a `Method`;
`Predecessor` wants a `Predecessor`; `Collected` wants `Pages`. artoo writes
what you give it, and a reviewer opening the define.html sees a blank where
you left one.

## ValueLevel — when one variable means several things

A `QVAL` in a SUPP dataset, or an `AVAL` that is a different measurement per
`PARAMCD`, needs one row per meaning. The columns are the Variables ones,
plus the one that matters most:

`Where Clause` says **which records this row describes**. In the newer
shape you write the condition itself; in the older shape you write an ID
and define the condition on the WhereClauses sheet.

The condition grammar, written out:

```
PARAMCD EQ ACTOT                              a single value
PARAMCD IN (ACITM01, ACITM02, ACITM03)        a set, in parentheses
LBTESTCD EQ HCT and LBSPEC EQ BLOOD           several conditions, ANDed
AVISIT EQ "Week 24"                           quote a value with a space
LBCAT EQ "CHEMISTRY, LOCAL"                   or with a comma
DM.COUNTRY EQ USA                             a variable in another dataset
```

Three rules worth knowing, because each has bitten someone:

- **Quotes are delimiters, not part of the value.** `AVISIT EQ "Week 24"`
  selects records whose AVISIT is the seven characters `Week 24`. You only
  need them when the value contains a space or a comma.
- **`IN` takes parentheses; `EQ` does not.** A comma inside an `EQ` value is
  part of the value.
- **Qualify a variable from another dataset**, as `DM.COUNTRY`. Without the
  prefix, the condition is read as naming a variable in the row's own
  dataset — a different question with a different answer.

## WhereClauses — the older shape only

One row per comparison; several rows sharing an `ID` are ANDed together.

| Column | Notes |
|---|---|
| `ID` | Referenced from the ValueLevel `Where Clause` cell. |
| `Dataset`, `Variable` | What is being tested. |
| `Comparator` | `EQ`, `NE`, `IN`, `NOTIN`, `LT`, `LE`, `GT`, `GE`. |
| `Value` | One value, or a comma-separated list for `IN` / `NOTIN`. |

If your workbook has no such sheet, write the condition in the ValueLevel
cell instead; artoo works out which shape you used from whether the sheet
is there.

## Codelists — one row per term

The list-level columns repeat on every term row. That looks redundant and is
how the format works.

| Column | Notes |
|---|---|
| `ID` | The codelist identifier, referenced from `Variables.Codelist`. |
| `Name` | The codelist name. **Required** by the schema. |
| `Data Type` | `text`, `integer`, `float`. Must match the variables that use it. |
| `NCI Codelist Code` | The C-code for the list, e.g. `C66731`. |
| `Order` | Position of the term. |
| `Term` | The submitted value. |
| `Decoded Value` | What it means. |
| `NCI Term Code` | The C-code for the term. |
| `Comment` | A **Comment ID**. |

## Dictionaries — MedDRA, WHODrug, ISO 3166

An external dictionary has no terms to list, so it gets its own sheet: `ID`,
`Name`, `Data Type`, `Dictionary`, `Version`. A variable points at one from
its **`Codelist`** column, exactly as it would an enumerated codelist —
there is one column for both, and artoo resolves against both.

## Methods, Comments, Documents

**Methods** — `ID`, `Name`, `Type` (`Computation`, `Imputation`), and
`Description`, the prose a reviewer reads. `Expression Context` and
`Expression Code` carry one formal expression: the language, and the code.

**Comments** — `ID` and `Description`. Every `Comment` cell elsewhere in the
workbook is an ID matching a row here. This is the single most common
workbook mistake: writing the comment text into the `Comment` cell of the
Variables sheet, where an ID belongs.

**Documents** — `ID`, `Title`, `Href`. The `Href` is a filename, and it is
also how artoo knows which document is the annotated CRF: **a filename
ending in `crf`** (`acrf.pdf`, `blankcrf.pdf`, `oncology-crf.pdf`) is the
annotated CRF, and everything else is a supplemental document. If your CRF
is called something else, say so with a `Role` column of `Annotated CRF`.

## Pages

A page reference means nothing without a document, so `Pages` is read
together with the annotated CRF above. Write them as `4 5`, `4, 5`, or a
range `4-5` — all three are understood, and a range is written out as a
first and last page rather than a list.

## Analysis Displays and Analysis Results

For ADaM specifications with analysis results metadata.

**Analysis Displays** — `ID`, `Title`, and the `Document` and `Pages` where
the display appears.

**Analysis Results** — `Display`, `ID`, `Description`, `Reason`, `Purpose`,
the documentation and the programming code. The column that carries the
selection is:

- newer shape: `Selection Criteria`, one bracket group per analysis
  dataset — `ADAE[AESER EQ Y] ADSL[SAFFL EQ Y]`, and `ADSL[]` for every
  record;
- older shape: an `Analysis Criteria` sheet, one row per analysis dataset
  with its own `Variables` and `Where Clause`.

`Variables` names the analysis variables, qualified by dataset when the
result spans more than one: `ADAE.AEBODSYS, ADAE.AEDECOD`.

## Fill it in, then check it

Reading tells you what artoo understood. Read the workbook back and look at
the slots before you write anything:

```{r}
spec <- read_spec(system.file("extdata", "adam-spec.xlsx", package = "artoo"))
spec
```

`validate_spec()` checks the spec against itself — every reference resolving,
every key sequence contiguous — without needing the data:

```{r}
validate_spec(spec)
```

Then write the define.xml. artoo validates the document against the CDISC
schemas before the file reaches its destination, so an invalid one never
overwrites a good file:

```{r}
define <- tempfile(fileext = ".xml")
write_spec(spec, define)
validate_define(define)
```

And `lint_define()` walks the reference graph the schema cannot see — an OID
in Define-XML is a plain string as far as the schema is concerned, so a
reference pointing at nothing is schema-valid:

```{r}
lint_define(define)
```

## What a workbook cannot carry

A workbook is an interchange format, not a lossless one. A spec that came
from a define.xml knows things no sheet has a column for — the OIDs the
sponsor chose, which standard each dataset follows, whether a codelist was
extended. artoo names them rather than dropping them quietly:

```{r}
#| warning: true
book2 <- tempfile(fileext = ".xlsx")
write_spec(adam_spec, book2)
```

Each line names the columns that sheet has no place for. Write a `.json`
path instead when you need the spec back exactly — that format is lossless
by construction, and it is the one to archive.

## Where to next

- [Specifications](specs.html) — the same workbook from the R side: read,
  inspect, repair, and write.
- [Conform & validate](conform.html) — apply the spec to data and see every
  finding.
- [Get started](artoo.html) — the whole round trip.
