---
title: "Authoring a specification workbook"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(artoo)
```

Most specifications are written in Excel, in a workbook of one sheet per
kind of metadata. artoo reads that workbook and writes the submission's
define.xml from it. This article is the reference for filling one in: what
each sheet is for, which cells artoo actually reads, and the handful of
places where a plausible-looking entry produces a document that is not what
you meant.

It is written for the person holding the workbook. If you want the R side —
inspecting a spec, repairing it, writing the define.xml — that is
[Specifications](specs.html).

## Start from a blank one

`write_template()` writes an empty workbook whose sheets and headers are
artoo's own reader, so nothing you fill in can be silently ignored:

```{r}
book <- tempfile(fileext = ".xlsx")
write_template(book)
readxl::excel_sheets(book)
```

Only **Datasets** and **Variables** are required. Every other sheet may be
left empty and artoo omits what it finds nothing in.

Two things to know before you start. A **column** artoo does not recognise
is kept and written back out, so a sponsor column of your own survives. A
**sheet** artoo does not recognise is not — add rows to the sheets above
rather than a sheet of your own.

## Two shapes, and how to tell them apart

Workbooks come in two generations, and artoo reads both. You will recognise
which one you have from three things:

| | Older | Newer |
|---|---|---|
| The study sheet is called | `Study` | `Define` |
| Conditions live | on a `WhereClauses` sheet | in the `ValueLevel` cell |
| The label column is called | `Description` | `Label` |

If your workbook came out of a validation tool, it is whichever that tool
writes. If you are starting fresh, use the template above — it is the newer
shape — and read on; anywhere the two differ, this article says so.

## Define / Study — the study sheet

Two columns, `Attribute` and `Value`, one row per attribute.

These five are the ones to fill in:

| Attribute | Why it matters |
|---|---|
| `StudyName` | The study's short name. |
| `StudyDescription` | One line describing the study. |
| `ProtocolName` | The protocol identifier. |
| `StandardName` | `SDTM-IG`, `ADaM-IG`, `SEND-IG`. Unhyphenated spellings are understood too. |
| `StandardVersion` | `3.4`, `1.1`, and so on. |

**`StandardName` and `StandardVersion` are the two that bite.** Define-XML
2.0 refuses to be written without them; 2.1, which artoo writes by default,
warns and carries on. Fill them in either way — everything downstream reads
better for it. Write the older version with
`write_spec(spec, path, version = "2.0")`.

A workbook artoo wrote from an existing define.xml carries more, because
that document had identifiers this sheet is where to keep:

```{r}
book3 <- tempfile(fileext = ".xlsx")
suppressWarnings(write_spec(adam_spec, book3))
readxl::read_excel(book3, sheet = "Define")
```

You do not have to supply any of those; artoo mints what a new document
needs. Leave them alone if they are there.

Anything else you put on this sheet is kept and written back out, so a
sponsor field of your own survives a round trip.

## Datasets — one row per dataset

`Dataset` and one of `Description` / `Label` are the minimum.

| Column | Notes |
|---|---|
| `Dataset` | The name, `DM`, `ADSL`. |
| `Description` or `Label` | The dataset label. |
| `Class` | `EVENTS`, `FINDINGS`, `BASIC DATA STRUCTURE`, ... |
| `Structure` | Prose: "One record per subject". **Required** — Define-XML has no default for it. |
| `Key Variables` | Space- or comma-separated. This is where a variable's key sequence comes from, so you do not fill that in twice. |
| `Purpose` | `Tabulation` or `Analysis`. Derived from the standard when blank. |
| `Repeating`, `Reference Data` | `Yes` / `No`. |
| `Domain` | The two-letter domain, for SDTM. Named in the not-submission-grade warning when blank. |
| `SubClass` | A subclass of `Class`, where the standard defines one. |
| `SAS Dataset Name` | When it differs from `Dataset`. Defaults to it. |
| `Has No Data` | `Yes` when the dataset is empty. Needs a `Comment` explaining why. |
| `Standard` | Which standard this dataset follows, when a study mixes versions. |
| `Comment` | A **Comment ID**, matching a row on the Comments sheet — not free text. |

## Variables — one row per variable per dataset

| Column | Notes |
|---|---|
| `Order` | Position within the dataset. |
| `Dataset`, `Variable` | Together they identify the row. |
| `Label` | The variable label. |
| `Data Type` | `text`, `integer`, `float`, `date`, `datetime`, `time`. |
| `Length` | Expected on text, integer and float. **Leave it blank on dates and times** — a validator flags one there. |
| `Significant Digits` | Only with `float`. |
| `Format` | The display format. |
| `Mandatory` | `Yes` / `No`. |
| `Codelist` | A Codelist ID **or a Dictionary ID** — one column serves both. |
| `Origin` | `Collected`, `Derived`, `Assigned`, `Protocol`, `Predecessor`, `Not Available`, `Other`. **The capitals matter.** |
| `Pages` | CRF page numbers. See *Pages* below. |
| `Method` | A Method ID, for a derived variable. |
| `Predecessor` | Where an ADaM variable came from, e.g. `DM.AGE`. |
| `Role` | `Identifier`, `Topic`, `Timing`, ... |
| `Origin Document` | Which document `Pages` refers to. Defaults to the annotated CRF. |
| `Source` | Who provided a collected value: `Investigator`, `Sponsor`, `Subject`, `Vendor`. |
| `Assigned Value` | The value, for an `Assigned` origin. |
| `SAS Field Name` | When it differs from `Variable`. |
| `Has No Data` | `Yes` when the variable is always empty. |
| `Comment` | A Comment ID. |

**Origin decides what else you must fill.** `Derived` wants a `Method`;
`Predecessor` wants a `Predecessor`; `Collected` wants `Pages`. artoo writes
what you give it, and a reviewer reading the rendered document sees a blank where you
left one.

## ValueLevel — when one variable means several things

A `QVAL` in a SUPP dataset, or an `AVAL` that is a different measurement per
`PARAMCD`, needs one row per meaning. The columns are the Variables ones,
plus the one that matters most:

`Where Clause` says **which records this row describes**. In the newer
shape you write the condition itself; in the older shape you write an ID
and define the condition on the WhereClauses sheet.

The condition grammar, written out:

```
PARAMCD EQ ACTOT                              a single value
PARAMCD IN (ACITM01, ACITM02, ACITM03)        a set, in parentheses
LBTESTCD EQ HCT and LBSPEC EQ BLOOD           several conditions, ANDed
AVISIT EQ "Week 24"                           quote a value with a space
LBCAT EQ "CHEMISTRY, LOCAL"                   or with a comma
DM.COUNTRY EQ USA                             a variable in another dataset
```

Four rules worth knowing:

- **A value runs to the end of the condition.** `AVISIT EQ Week 24` already
  means the seven characters `Week 24`, and `LBCAT EQ CHEMISTRY, LOCAL`
  already means the sixteen. You do not need quotes for a space or a comma
  after `EQ`.
- **Double quotes are stripped when you use them**, so writing
  `AVISIT EQ "Week 24"` gets the same answer. Single quotes are not: they
  stay in the value.
- **The one place quotes are load-bearing is inside a set.** `IN` splits on
  commas, so a member that contains one has to be quoted:
  `PARAMCD IN ("A, B", C)` is two members, not three. Parentheses round a
  set are optional.
- **Qualify a variable from another dataset**, as `DM.COUNTRY`. Without the
  prefix, the condition is read as naming a variable in the row's own
  dataset — a different question with a different answer.

**`or` is not a general operator.** Conditions are ANDed. artoo folds an
`or` between values of the *same* variable into a set — `SEX EQ M or SEX EQ
F` becomes `SEX IN (M, F)` — and refuses one across different variables,
because Define-XML has no way to say it. Split those into separate
value-level rows.

**One thing cannot be written inline at all:** a value containing the word
`and` or `or`, because that is what separates conditions. `ATPT EQ 4 hours
and 30 minutes` cannot be parsed. Use a WhereClauses sheet, which needs no
parsing.

## WhereClauses — the older shape only

One row per comparison; several rows sharing an `ID` are ANDed together.

| Column | Notes |
|---|---|
| `ID` | Referenced from the ValueLevel `Where Clause` cell. |
| `Dataset`, `Variable` | What is being tested. |
| `Comparator` | `EQ`, `NE`, `IN`, `NOTIN`, `LT`, `LE`, `GT`, `GE`. |
| `Value` | One value, or a comma-separated list for `IN` / `NOTIN`. |
| `Comment` | A Comment ID, when the condition needs explaining. |

If your workbook has no such sheet, write the condition in the ValueLevel
cell instead; artoo works out which shape you used from whether the sheet
is there.

## Codelists — one row per term

The list-level columns repeat on every term row. That looks redundant and is
how the format works.

| Column | Notes |
|---|---|
| `ID` | The codelist identifier, referenced from `Variables.Codelist`. |
| `Name` | The codelist name. Fill it in — artoo falls back to the ID, which is not a name. |
| `Data Type` | `text`, `integer`, `float`. Should match the variables that use it. |
| `NCI Codelist Code` | The C-code for the list, e.g. `C66731`. |
| `Order` | Position of the term. |
| `Term` | The submitted value. |
| `Decoded Value` | What it means. |
| `NCI Term Code` | The C-code for the term. |
| `Rank` | The term's rank, where the terminology defines one. |
| `SAS Format Name` | The format name, `$SEX`. |
| `Comment` | A **Comment ID**. |

## Dictionaries — MedDRA, WHODrug, ISO 3166

An external dictionary has no terms to list, so it gets its own sheet: `ID`,
`Name`, `Data Type`, `Dictionary`, `Version`. A variable points at one from
its **`Codelist`** column, exactly as it would an enumerated codelist —
there is one column for both, and artoo resolves against both.

## Standards

Define-XML 2.1 records every standard the study uses, not just the one the
study sheet names: the implementation guide, the controlled terminology
releases, any prior versions still in play. One row each — `ID`, `Name`,
`Type` (`IG`, `CT`, `MODEL`), `Version`, `Status`, `Publishing Set` — and
the Datasets sheet's `Standard` column points at them.

Leave it empty and artoo writes one row from the study sheet's
`StandardName` and `StandardVersion`, which is enough for a document that
follows a single standard.

## Methods, Comments, Documents

**Methods** — `ID`, `Name`, `Type` (`Computation`, `Imputation`,
`Transpose`, `Other`), and `Description`, the prose a reviewer reads.
`Expression Context` and `Expression Code` carry one formal expression: the
language, and the code. `Document` and `Pages` point at where the method is
written up.

**Comments** — `ID`, `Description`, and a `Document` and `Pages` of their
own. Every `Comment` cell elsewhere in the workbook is an ID matching a row
here. This is the single most common workbook mistake: writing the comment
text into the `Comment` cell of the Variables sheet, where an ID belongs.
artoo reports it as an unresolved reference rather than guessing.

**Documents** — `ID`, `Title`, `Href`. The `Href` is a filename, and it is
also how artoo knows which document is the annotated CRF: **a filename
ending in `crf`** (`acrf.pdf`, `blankcrf.pdf`, `oncology-crf.pdf`) is the
annotated CRF, and everything else is a supplemental document. If your CRF
is called something else, say so with a `Role` column of `Annotated CRF`.

## Pages

A page reference means nothing without a document. Write pages as `4 5`,
`4, 5`, or a range `4-5`: all three are understood, commas are normalised
to the spaces Define-XML wants, and a range is written out as a first and
last page rather than a list.

**Which document the pages belong to** comes from one of two places. Set
`Origin Document` on the row to name a specific one; leave it blank and a
`Collected` origin's pages default to the annotated CRF. Methods and
Comments have their own `Document` column for the same reason. If a row has
pages and there is no document to attach them to, artoo refuses rather than
attaching them to a document nobody named.

## Analysis Displays and Analysis Results

For ADaM specifications with analysis results metadata.

**Analysis Displays** — `ID`, `Title`, and the `Document` and `Pages` where
the display appears.

**Analysis Results** — `Display`, `ID`, `Description`, `Reason`, `Purpose`,
the documentation and the programming code. The column that carries the
selection is:

- newer shape: `Selection Criteria`, one bracket group per analysis
  dataset — `ADAE[AESER EQ Y] ADSL[SAFFL EQ Y]`, and `ADSL[]` for every
  record;
- older shape: an `Analysis Criteria` sheet, one row per analysis dataset,
  keyed by `Display` and `Result`, with its own `Variables` and
  `Where Clause` — the same condition grammar as ValueLevel.

`Variables` names the analysis variables, qualified by dataset when the
result spans more than one: `ADAE.AEBODSYS, ADAE.AEDECOD`. `Documentation
Refs` and `Programming Document` name a Document ID each. `Join Comment` is
a Comment ID explaining how several analysis datasets relate.

## Fill it in, then check it

Reading tells you what artoo understood. Read the workbook back and look at
the slots before you write anything:

```{r}
spec <- read_spec(system.file("extdata", "adam-spec.xlsx", package = "artoo"))
spec
```

`validate_spec()` checks the spec against itself — every reference
resolving, every key sequence contiguous — without needing the data. It
reports notes as well as errors, so expect to see some on a real spec:

```{r}
validate_spec(spec)
```

Then write the define.xml. artoo validates the document against the CDISC
schemas before the file reaches its destination, so an invalid one never
overwrites a good file:

```{r}
define <- tempfile(fileext = ".xml")
write_spec(spec, define)
validate_define(define)
```

Pass `html = TRUE` to also render the document through the CDISC
stylesheet. One quirk of that stylesheet to not chase: every 2.1 dataset
heading ends `... - ` with the class blank, because the heading reads the
2.0 class *attribute* and 2.1 records the class as an *element* — CDISC's
own published 2.1 examples render the same bare dash, and the class does
appear in the Datasets summary table.

And `lint_define()` walks the reference graph the schema cannot see — an OID
in Define-XML is a plain string as far as the schema is concerned, so a
reference pointing at nothing is schema-valid:

```{r}
lint_define(define)
```

Read the severities, not the count. An **error** is a reference pointing at
nothing, and it means the document is wrong. A **warning** is usually an
orphan — something defined that nothing uses — which a valid document may
well have, and this bundled spec does: it is scoped to two datasets and
still names documents and standards the others used. Nothing here is
dangling, which is the line that matters.

## What a workbook cannot carry

A workbook is an interchange format, not a lossless one. A spec that came
from a define.xml knows things no sheet has a column for — the OIDs the
sponsor chose, which standard each dataset follows, whether a codelist was
extended. artoo names them rather than dropping them quietly:

```{r}
#| warning: true
book2 <- tempfile(fileext = ".xlsx")
write_spec(adam_spec, book2)
```

Each line names the columns that sheet has no place for. Write a `.json`
path instead when you need the spec back exactly — that format is lossless
by construction, and it is the one to archive.

## Where to next

- [Specifications](specs.html) — the same workbook from the R side: read,
  inspect, repair, and write.
- [Conform & validate](conform.html) — apply the spec to data and see every
  finding.
- [Get started](artoo.html) — the whole round trip.
